---
title: "Cells_05"
output: html_document
date: "2025-06-25"
---
# RNA triplicates pooled within each experiment: 
# Experiment 1 -> Control 1 + Exposed 1
# Experiment 2 -> Control 2 + Exposed 2
# Experiment 3 -> Control 3 + Exposed 3 
# As each experiment contributed exactly this will be a paired design, where pairing is by experiment (rather than donor; see placenta transcriptomic work)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install DESeq2
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")

#install tidyverse (ggplot2 is in tidyverse)
#install.packages("tidyverse")

```
# Package Setup
```{r}
library( "DESeq2" )
library(tidyverse)
library(ggplot2)
library(here)   # builds file paths relative to the project root (the .Rproj folder)
```


```{r}
# Point at the input folder relative to the project root -- no setwd() needed.
# here() resolves to wherever this project lives, so it works on any computer.
cells_dir <- here("data", "cells")

# cells_1_5um

cells_1_5um_count_data <- read.csv(file.path(cells_dir, 'N1_5um_counts.tabular'), header = TRUE, sep = "\t")
colnames(cells_1_5um_count_data)[2] = "cells_1_5um"

# cells_1_Control

cells_1_Control_count_data <- read.csv(file.path(cells_dir, 'N1_Control_counts.tabular'), header = TRUE, sep = "\t")
colnames(cells_1_Control_count_data)[2] = "cells_1_Control"

# cells_2_5um

cells_2_5um_count_data <- read.csv(file.path(cells_dir, 'N2_5um_counts.tabular'), header = TRUE, sep = "\t")
colnames(cells_2_5um_count_data)[2] = "cells_2_5um"

# cells_1_Control

cells_2_Control_count_data <- read.csv(file.path(cells_dir, 'N2_Control_counts.tabular'), header = TRUE, sep = "\t")
colnames(cells_2_Control_count_data)[2] = "cells_2_Control"

# cells_3_5um

cells_3_5um_count_data <- read.csv(file.path(cells_dir, 'N3_5um_counts.tabular'), header = TRUE, sep = "\t")
colnames(cells_3_5um_count_data)[2] = "cells_3_5um"

# cells_1_Control

cells_3_Control_count_data <- read.csv(file.path(cells_dir, 'N3_Control_counts.tabular'), header = TRUE, sep = "\t")
colnames(cells_3_Control_count_data)[2] = "cells_3_Control"

```

# Join all count data into one data frame by full_join & remove NA values with 0 
```{r}

#put all data frames into list 

df_list <- list(cells_3_Control_count_data, cells_3_5um_count_data, cells_2_Control_count_data, cells_2_5um_count_data, cells_1_Control_count_data, cells_1_5um_count_data)

library(purrr)

# merge all data frames into list 
count_data <- purrr::reduce(df_list, dplyr::full_join, by= 'Geneid')


# make the Geneid the row id 
count_data <- count_data %>% remove_rownames %>% column_to_rownames(var="Geneid")

cells_5um_count_data_file_path <- here("output", "cells", "cells_count_data.csv")
write.csv(as.data.frame(count_data), file = cells_5um_count_data_file_path, row.names = TRUE)

```

# Create Metadata
## Goal: Comparing "control" vs. "treated" across three experiments
```{r}
SampleName <- c(colnames(count_data))
treatment <- c("control", "treated", "control", "treated", "control", "treated")
experiment <- c("exp3","exp3","exp2", "exp2","exp1","exp1")

meta_data <- data.frame(SampleName, treatment, experiment)
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="SampleName")
meta_data$treatment <- as.factor(meta_data$treatment)
meta_data$experiment <- as.factor(meta_data$experiment)

head(meta_data)

```
# Double check that the name in the Columns in the Counts Matrix = Nmaes in the Meta Data Frame 

```{r}

all(colnames(count_data) %in% rownames(meta_data))

all(colnames(count_data) == rownames(meta_data))
```

# Create DESeqDataSet object

# Model gene expression as a function treatment, while controlling for experiment (blocking plate to plate variation) 
```{r}
design_formula <- ~ experiment + treatment 
dds <- DESeqDataSetFromMatrix(countData = count_data,
                              colData = meta_data,
                              design = design_formula)
#shows object elements
dim(dds)
```
# Pre-filtering: by removing rows in which there are very few reads
# filter any counts less than 200 
```{r}
keep <- rowSums(counts(dds)) >= 200
dds <- dds[keep,]
dds
head(dds)

resultsNames(dds)
```

# Relevel Reference Factors 
```{r}
dds$treatment <- relevel(dds$treatment, ref = 'control')
dds$experiment <- relevel(dds$experiment, ref = 'exp1')

levels(dds$treatment) #control should be first 
levels(dds$experiment) #"exp1" should be first 
```

# PCA plot
```{r}
library(DESeq2)
library(ggplot2)

# Transform using VST 
vsd.all <- vst(dds, blind = TRUE)

# Relabel factors BEFORE plotting
colData(vsd.all)$treatment <- factor(
  colData(vsd.all)$treatment,
  levels = c("control", "treated"),
  labels = c("Control", "5 µM")
)
colData(vsd.all)$experiment <- factor(
  colData(vsd.all)$experiment,
  levels = c("exp1", "exp2", "exp3"),
  labels = c("N1", "N2", "N3")
)

# Verify metadata looks correct BEFORE plotting
head(colData(vsd.all))

# Extract PCA data for ggplot
pcaData <- plotPCA(
  vsd.all,
  intgroup = c("treatment", "experiment"),
  returnData = TRUE
)

# Extract percent variance
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Check pcaData looks right
head(pcaData)

# Build Plots
p1 <- ggplot(pcaData, aes(PC1, PC2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(
    values = c(
      "Control" = "#1b9e77",
      "5 µM"    = "#d95f02"
    )
  ) +
  labs(
    title = "HTR-8/SVneo: Control vs. Treatment",
    x     = paste0("PC1: ", percentVar[1], "% variance"),
    y     = paste0("PC2: ", percentVar[2], "% variance"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(
    plot.title   = element_text(hjust = 0.5, size = 11),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text.x  = element_text(size = 9),
    axis.text.y  = element_text(size = 9),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9),
    plot.margin  = margin(t = 10, r = 15,
                          b = 10, l = 10,
                          unit = "pt")
  )

# Save as TIFF
ggsave(
  filename    = "Figure1a_PCA_cells.tiff",
  plot        = p1,
  path        = here("output", "figures"),
  width       = 4.5,
  height      = 3.5,
  units       = "in",
  dpi         = 1200,
  device      = "tiff",
  compression = "lzw"
)

p2 <- ggplot(pcaData, aes(PC1, PC2, color = experiment)) +
  geom_point(size = 4) +
  scale_color_manual(values = c(
    "N1" = "#66c2a5",
    "N2" = "#fc8d62",
    "N3" = "#8da0cb"
  )) +
  labs(
    title = "HTR-8/SVneo: By Experiment",
    x     = paste0("PC1: ", percentVar[1], "% variance"),
    y     = paste0("PC2: ", percentVar[2], "% variance"),
    color = "Experiment"
  ) +
  theme_bw() +
  theme(
    plot.title   = element_text(hjust = 0.5, size = 11),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text.x  = element_text(size = 9),
    axis.text.y  = element_text(size = 9),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9),
    plot.margin  = margin(t = 10, r = 15,
                          b = 10, l = 10,
                          unit = "pt")
  )

# Save p2
ggsave(
  filename    = "Figure1b_PCA_cells.tiff",
  plot        = p2,
  path        = here("output", "figures"),
  width       = 4.5,
  height      = 3.5,
  units       = "in",
  dpi         = 1200,
  device      = "tiff",
  compression = "lzw"
)

# Preview both in RStudio
print(p1)
print(p2)
```

# Test 
```{r}
dds <- DESeq(dds) #testing

res.all <- results(dds, independentFiltering = FALSE)  # Extract results table

mcols(res.all)
summary(res.all,alpha=0.05) #viewing summary of DE genes

res.all$gene <- rownames(res.all) #adding actual columnn with gene names
res.all <- res.all [,c('gene','log2FoldChange','baseMean','lfcSE', 'stat','pvalue','padj')] #reordering columns
rownames(res.all)<-NULL #removing non-column row names (genes)
dim(res.all) #checking dataframe
head(res.all)#checking dataframe

log2foldchange_cells_results_file_path <- here("output", "cells", "results_cells_5.csv")
write.csv(as.data.frame(res.all), file = log2foldchange_cells_results_file_path, row.names = TRUE)

resultsNames(dds)
```
# Checking distribution of p-values 
```{r}
hist(res.all$pvalue, breaks = 50,
     main = "P-value Distribution",
     xlab = "p-value",
     col = "steelblue")

```

# Save as .rnk file for Pathway Analysis in GSEA 
```{r}
ranked_list <- as.data.frame(res.all) %>%   # DESeqResults -> data.frame so dplyr verbs work
  filter(!is.na(log2FoldChange)) %>%
  filter(!is.na(gene)) %>%
  filter(gene != "") %>%
  group_by(gene) %>%
  slice_max(
    order_by  = abs(log2FoldChange),
    n         = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  arrange(desc(log2FoldChange)) %>%
  select(gene, log2FoldChange)


head(ranked_list)
tail(ranked_list)
cat("Total genes:", nrow(ranked_list), "\n")


write.table(
  ranked_list,
  # was on the Desktop; now written inside the project so it travels with the repo
  file      = here("output", "cells", "cells_5_preranked.rnk"),
  sep       = "\t",          # tab separated
  row.names = FALSE,         # no row numbers
  col.names = FALSE,         # no header
  quote     = FALSE,         # no quotes
  fileEncoding = "UTF-8"     # standard encoding GSEA can read
)

cat("RNK file saved!\n")

# Step 4 - verify file looks correct
# open in Notepad NOT Excel to check
verify <- read.table(
  here("output", "cells", "cells_5_preranked.rnk"),
  sep = "\t",
  stringsAsFactors = FALSE
)
```

# New Threshold Criteria: Getting the # of DEG which meet both criteria 

```{r}
#install.packages("DT")
library(DT)
# Condition thresholds 
padj_cutoff <- 0.05 
lfc_cutoff <- 0.585 #log2(1.5) = 0.585

# Both padj < 0.05 AND absolute log2FC >= 0.585 

both <- res.all[res.all$padj < padj_cutoff & abs(res.all$log2FoldChange) >= lfc_cutoff, ] 
both <- as.data.frame(both)
dim(both) 
options(max.print = 1000)
print(both)

DT:: datatable(both)

# Only padj significant 
padj_only <- res.all[res.all$padj < padj_cutoff & abs(res.all$log2FoldChange) < lfc_cutoff, ] 
padj_only <- as.data.frame(padj_only)
dim(padj_only) 
options(max.print = 1000)
print(padj_only)

DT:: datatable(padj_only)

# Only log2FC significant 
lfc_only <- res.all[res.all$padj >= padj_cutoff & abs(res.all$log2FoldChange) >= lfc_cutoff, ]
lfc_only <- as.data.frame(lfc_only)
dim(lfc_only) 
options(max.print = 1000)
print(lfc_only)

DT:: datatable(lfc_only)

#install.packages("openxlsx")
library(openxlsx)
write.xlsx(list(
  "Both" = both,
  "Adjusted P-Value Only" = padj_only,
  "LogFC Only" = lfc_only
), file = here("output", "cells", "cells_5_gene_counts_threshold_L(1.5)FC.xlsx"))

```

# Identifying Up and Down Regulated Genes 
```{r}
# add an additional column that identifies a gene as unregulated, downregulated, or unchanged
# note the choice of pvalue and log2(1.5)FoldChange cutoff. 
res.all <- as.data.frame(res.all)
res.all <- res.all %>%
  mutate(
  Expression = case_when
  (log2FoldChange >= log2(1.5) & padj <= 0.05 ~ "Up-regulated",
  log2FoldChange <= -log2(1.5) & padj <= 0.05 ~ "Down-regulated",
  TRUE ~ "Unchanged")
  )
head(res.all)
```
# Which Genes are the Top 10 Most Variable?
```{r}
top <- 10
# we are getting the top 10 up and down regulated genes by filtering the column Up-regulated and Down-regulated and sorting by the adjusted p-value. 

top_genes <- bind_rows(
  res.all %>%
  filter(Expression == 'Up-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top),
  res.all %>%
  filter(Expression == 'Down-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top)
  )

# create a data frame just holding the top 10 genes

Top_Hits = head(arrange(res.all,pvalue),10)
Top_Hits
```


```{r}
colnames(res.all)
```
# Volcano Plot (Code developed with assistance from Claude (claude-sonnet-4-6, Anthropic, 2026))
# Dimensions tailored to your data range
# Max -log10(padj) = 8.9
# Max |LFC| = 1.37
```{r}
head(res.all)

library(ggplot2)
library(ggrepel)
library(dplyr)

# STEP 1 — Prepare volcano data
# Dimensions tailored to your data range
# Max -log10(padj) = 8.9
# Max |LFC| = 1.37
# ============================================

volcano_data <- res.all %>%
  filter(!is.na(pvalue)) %>%
  filter(!is.na(padj)) %>%
  filter(!is.na(log2FoldChange)) %>%
  mutate(
    neg_log10_padj = pmin(-log10(padj), 10),

    significance = case_when(
      padj < 0.05 & log2FoldChange >  log2(1.5) ~ "Upregulated",
      padj < 0.05 & log2FoldChange < -log2(1.5) ~ "Downregulated",
      padj < 0.05                                ~ "Significant (|LFC| < 0.585)",
      TRUE                                       ~ "Not Significant"
    ),

    significance = factor(
      significance,
      levels = c(
        "Upregulated",
        "Downregulated",
        "Significant (|LFC| < 0.585)",
        "Not Significant"
      )
    )
  )

cat("Gene counts by category:\n")
print(table(volcano_data$significance))

top_up <- volcano_data %>%
  filter(significance == "Upregulated") %>%
  arrange(padj) %>%
  head(10)

top_down <- volcano_data %>%
  filter(significance == "Downregulated") %>%
  arrange(padj) %>%
  head(10)

top_sig <- volcano_data %>%
  filter(significance == "Significant (|LFC| < 0.585)") %>%
  arrange(padj) %>%
  head(10)

if (nrow(top_up) > 0 | nrow(top_down) > 0) {
  genes_to_label <- bind_rows(top_up, top_down)
  cat("\nLabeling up and downregulated genes\n")
} else {
  genes_to_label <- top_sig
  cat("\nLabeling top significant genes\n")
}

cat("Genes labeled:", nrow(genes_to_label), "\n")
print(genes_to_label$gene)

teal_colors <- c(
  "Upregulated"              = "#004c4c",
  "Downregulated"            = "#008080",
  "Significant (|LFC| < 0.585)" = "#5ba8a8",
  "Not Significant"          = "#d4eaea"
)

volcano_plot <- ggplot(
  volcano_data,
  aes(x     = log2FoldChange,
      y     = neg_log10_padj,
      color = significance)
) +
  geom_point(
    data  = volcano_data %>% filter(significance == "Not Significant"),
    alpha = 0.4,
    size  = 1.5
  ) +
  geom_point(
    data  = volcano_data %>% filter(significance != "Not Significant"),
    alpha = 0.9,
    size  = 2.0
  ) +
  scale_color_manual(
    values = teal_colors,
    name   = "Expression",
    labels = c(
      "Upregulated" = paste0(
        "Upregulated (padj<0.05, LFC>0.585)  n=",
        sum(volcano_data$significance == "Upregulated", na.rm = TRUE)
      ),
      "Downregulated" = paste0(
        "Downregulated (padj<0.05, LFC<\u22120.585)  n=",
        sum(volcano_data$significance == "Downregulated", na.rm = TRUE)
      ),
      "Significant (|LFC| < 0.585)" = paste0(
        "Significant (padj<0.05, |LFC|<0.585)  n=",
        sum(volcano_data$significance == "Significant (|LFC| < 0.585)", na.rm = TRUE)
      ),
      "Not Significant" = paste0(
        "Not Significant  n=",
        sum(volcano_data$significance == "Not Significant", na.rm = TRUE)
      )
    )
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype   = "dashed",
    color      = "black",
    linewidth  = 0.5
  ) +
  geom_vline(
    xintercept = c(-log2(1.5), log2(1.5)),
    linetype   = "dashed",
    color      = "black",
    linewidth  = 0.5
  ) +
  geom_label_repel(
    data          = genes_to_label,
    aes(label     = gene),
    size          = 2.8,
    max.overlaps  = 20,
    box.padding   = 0.5,
    point.padding = 0.3,
    segment.color = "grey40",
    segment.size  = 0.3,
    show.legend   = FALSE,
    color         = "black",
    fill          = "white",
    label.size    = 0.25
  ) +
  labs(
    title    = "HTR-8/SVneo: Control vs. 5 µM PFBS",
    subtitle = paste0(
      "Upregulated: ",
      sum(volcano_data$significance == "Upregulated", na.rm = TRUE),
      "  |  Downregulated: ",
      sum(volcano_data$significance == "Downregulated", na.rm = TRUE),
      "  |  Significant (|LFC|<0.585): ",
      sum(volcano_data$significance == "Significant (|LFC| < 0.585)", na.rm = TRUE)
    ),
    x = expression(Log[2]~Fold~Change),
    y = expression(-Log[10]~adjusted~italic(p)~value)
  ) +
  scale_x_continuous(
    limits = c(-2, 2),
    breaks = seq(-2, 2, by = 0.5)
  ) +
  scale_y_continuous(
    limits = c(0, 10),
    breaks = seq(0, 10, by = 2)
  ) +
  theme_bw() +
  theme(
    plot.title       = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle    = element_text(hjust = 0.5, size = 9, color = "grey40"),
    axis.title.x     = element_text(size = 11),
    axis.title.y     = element_text(size = 11),
    axis.text.x      = element_text(size = 9),
    axis.text.y      = element_text(size = 9),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.text      = element_text(size = 8),
    legend.position  = "bottom",
    legend.direction = "vertical",
    plot.margin      = margin(t = 10, r = 15, b = 10, l = 10, unit = "pt")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 3, alpha = 1))
  )
ggsave(
  filename    = "New_Threshold_Figure_Volcano_Cells_5uM.tiff",
  plot        = volcano_plot,
  path        = here("output", "figures"),
  width       = 6,
  height      = 7,
  units       = "in",
  dpi         = 1200,
  device      = "tiff",
  compression = "lzw"
)

print(volcano_plot)
```

