---
title: "Comparisons & Correlations & Heatmap"
output: html_document
date: "2026-01-30"
---
# Comparison between Models; all genes of the cells_5uM and tissue_5uM

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(here)
cells_5uM_all_genes <- read_csv(here("output", "comparisons", "cells_5_results.csv"))
tissue_5uM_all_genes <- read_csv(here("output", "comparisons", "results_third_5.csv"))
```

```{r}

####cells_5uM vs tissue_5uM####
df_list <- list(cells_5uM_all_genes,tissue_5uM_all_genes)

# merge all data frames in list
df_list_next <- df_list %>% purrr::reduce(full_join, by='gene')

df_list_next <-df_list_next[complete.cases(df_list_next), ] #removes NA's#=
df_list_next<- na.omit(df_list_next)
dim(df_list_next)

cor.test(df_list_next$log2FoldChange.x,df_list_next$log2FoldChange.y)

pdf(file = here("output", "comparisons", "correlation_models.pdf"))

smoothScatter(df_list_next$log2FoldChange.x~df_list_next$log2FoldChange.y, colramp=colorRampPalette(c("white","violet")), cex=3,  xlim = c(-5, 5), ylim = c(-5, 5), main="0 (Control) vs. 5uM PFBS", xlab="Log2Fold Change (Cells-All Genes)", ylab ="Log2Fold Change (Explants-All Genes")

abline(v = 0, col="gray40")
abline(h = 0, col="gray40")
abline(lm (df_list_next$log2FoldChange.x~df_list_next$log2FoldChange.y), col="red",lwd=c(2,3))
dev.off()

```

# Hypergeometric Test between NFKB Gene Targets and Dose Response Gene List (PFBS_gene_listd)


```{r}
#Loading Packages & Dataframes 
library(readr)
PFBS_gene_list <- read_csv(here("output", "comparisons", "shared_placenta_5um_20um_801_DEGs.csv"))
NFKB_gene_target_list <- read_csv(here("output", "comparisons", "NFKB.csv"))
#install.packages("readxl")
library(readxl)
```

# Comparison between NFKB Gene Targets all genes of the cells_5uM and NFKB Gene Targets and all genes tissue_5uM

```{r}
#First create dataframes with genes in both models and NFKB list
#cells_NFKB
intersect(cells_5uM_all_genes$gene, NFKB_gene_target_list$gene) #which gene names in common between 2 groups
length (intersect(cells_5uM_all_genes$gene, NFKB_gene_target_list$gene)) # 145 # number of genes in common between each group
#Create Dataframe
cells_NFKB <- semi_join(cells_5uM_all_genes, NFKB_gene_target_list, by = "gene")
print(cells_NFKB)

#tissue_NFKB
intersect(tissue_5uM_all_genes$gene, NFKB_gene_target_list$gene) #which gene names in common between 2 groups
length (intersect(tissue_5uM_all_genes$gene, NFKB_gene_target_list$gene)) # 214 # number of genes in common between each group
#Create Dataframe
tissue_NFKB <- semi_join(tissue_5uM_all_genes, NFKB_gene_target_list, by = "gene")
print(tissue_NFKB)

```
Intersect separate overlaps of all genes with NFKB genes and THEN run a correlation 
```{r}
####cells_NFKB vs tissue_NFKB####

df_list_2 <-list(cells_NFKB,tissue_NFKB)
head(cells_NFKB)
head(tissue_NFKB)
```

```{r}
#merge all data frames into list
df_list_next2 <- df_list_2 %>% purrr::reduce(full_join, by='gene')

df_list_next2 <-df_list_next2[complete.cases(df_list_next2),] #removes NAs
df_list_next2 <- na.omit(df_list_next2)
dim(df_list_next2)
```


```{r}
cor.test(df_list_next2$log2FoldChange.x,df_list_next2$log2FoldChange.y)

pdf(file = here("output", "comparisons", "correlation_NFKB.pdf"))

smoothScatter(df_list_next2$log2FoldChange.x~df_list_next2$log2FoldChange.y, colramp=colorRampPalette(c("white","violet")), cex=3,  xlim = c(-5, 5), ylim = c(-5, 5), main="0 (Control) vs. 5uM PFBS", xlab="Log2FoldChange (Cells-NFKB Genes)", ylab ="Log2FoldChange (Explant-NFKB Genes)")
abline(v = 0, col="gray40")
abline(h = 0, col="gray40")
abline(lm (df_list_next2$log2FoldChange.x~df_list_next2$log2FoldChange.y), col="red",lwd=c(2,3))
dev.off()

```
# Determining the Number of Genes that Intersect between the Two Gene Lists: PFBS_gene_list & NFKB gene targets

```{r}
intersect(PFBS_gene_list$gene, NFKB_gene_target_list$gene) #which gene names in common between 2 groups
length (intersect(PFBS_gene_list$gene, NFKB_gene_target_list$gene)) # 46 # number of genes in common between each group

setdiff(PFBS_gene_list$gene, NFKB_gene_target_list$gene) #which genes names are in the first parameter but not second, flip parameters to get names for other parameter
length (setdiff(PFBS_gene_list$gene, NFKB_gene_target_list$gene)) # 755

setdiff(NFKB_gene_target_list$gene, PFBS_gene_list$gene)#which genes names are in the second parameter but not first, flip parameters to get names for other parameter
length (setdiff(NFKB_gene_target_list$gene, PFBS_gene_list$gene)) # 256

```
q = For NFKB only: intersect= 46
m = Number of DE gene FDR<0.05 and FC>1]= 801 
n = Total number of filtered genes minus DE expressed genes; the total number of genes in the background universe (genes tested in DESeq2 after filtering)= 14439 - 801 = 13638 
k = Number of genes in NFKB gene set= 302

#q=intersect; m=#of signficant DE genes; n=[total # filterd genes-#of DE expressed genes] k=# of NFKB only genes
```{r}
## Hypergeometric Testing with Phyper
# Is an overlap between my PFBS DEG set and the Preeclampsia gene set greater than expected under random sampling?

phyper(46, 801, 13638, 302, lower.tail= FALSE)

#p=  1.127607e-10

```

# Heatmap of all comparisons (N=5) 
```{r}
library(readxl)
library(pheatmap)
library(RColorBrewer)

wb_path <- here("output", "cells", "Total_Differential_Gene_Expression_Results .xlsx")

# Sheet names mapped to readable labels
sheets <- c(
  "Cells 5µm"    = "cells_5um",
  "Male 5µm"     = "male_5um",
  "Male 20µm"    = "male_20um",
  "Female 5µm"   = "female_5um",
  "Female 20µm"  = "female_20um"
)

# Read all sheets
de_list <- lapply(sheets, function(s) {
  suppressWarnings(
    read_excel(wb_path, sheet = s,
               col_types = c("skip", "text", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric"))
  )
})

# Collect genes significant (padj < 0.05) in at least one comparison
# Collect all significant genes
sig_genes <- unique(unlist(lapply(de_list, function(df) {
  df$gene[!is.na(df$padj) & df$padj < 0.05]
})))

# Build full log2FC matrix for significant genes
lfc_matrix <- sapply(de_list, function(df) {
  df$log2FoldChange[match(sig_genes, df$gene)]
})
rownames(lfc_matrix) <- sig_genes
colnames(lfc_matrix) <- names(sheets)

# For each gene, find its most extreme LFC across any comparison
max_lfc_per_gene <- apply(lfc_matrix, 1, function(x) max(x, na.rm = TRUE))
min_lfc_per_gene <- apply(lfc_matrix, 1, function(x) min(x, na.rm = TRUE))

# Top 50 upregulated and top 50 downregulated
top_up   <- names(sort(max_lfc_per_gene, decreasing = TRUE))[1:50]
top_down <- names(sort(min_lfc_per_gene, decreasing = FALSE))[1:50]
selected_genes <- unique(c(top_up, top_down))

cat("Genes in heatmap:", length(selected_genes), "\n")  # up to 100 if no overlap

# Subset and replace NAs with 0 for clustering
lfc_sub <- lfc_matrix[selected_genes, ]
lfc_sub[is.na(lfc_sub)] <- 0

# Color palette and breaks
max_val <- max(abs(lfc_sub), na.rm = TRUE)
breaks  <- seq(-max_val, max_val, length.out = 101)
colors  <- colorRampPalette(rev(brewer.pal(11, "RdBu")))(100)

pheatmap(
  lfc_sub,
  color         = colors,
  breaks        = breaks,
  cluster_rows  = TRUE,
  cluster_cols  = FALSE,
  show_rownames = TRUE,
  fontsize_row  = 6,
  fontsize_col  = 10,
  main          = "Top 50 Up- and Downregulated Genes (log2FC, padj < 0.05)",
  filename      = here("output", "figures", "heatmap_top50_up_down.tiff"),
  width  = 10,
  height = 14,
  res    = 300
)
```