---
title: "GSEA Sex Comparison - Placenta 20uM PFBS"
output: html_document
date: "2026-05-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(forcats)
library(ggVennDiagram)   # install.packages("ggVennDiagram") if needed
library(UpSetR)          # install.packages("UpSetR") if needed
```

```{r}
# ============================================================
# STEP 1 — Set paths to your GSEA output folders
# UPDATE THESE to match your actual folder names
# Each path should point to the folder GSEA created
# (the one containing gsea_report_for_na_pos/neg .tsv files)
# ============================================================

# --- HALLMARK ---
male_hallmark_path   <- "C:/Users/Maggie Drelichman/Thesis/data/gsea/PLACENTA_MALE_HALLMARK"
female_hallmark_path <- "C:/Users/Maggie Drelichman/Thesis/data/gsea/PLACENTA_FEMALE_HALLMARK"

# --- REACTOME (add when ready) ---
# male_reactome_path   <- "C:/Users/Maggie Drelichman/Thesis/data/gsea/PLACENTA_MALE_REACTOME"
# female_reactome_path <- "C:/Users/Maggie Drelichman/Thesis/data/gsea/PLACENTA_FEMALE_REACTOME"

# --- GOBP (add when ready) ---
# male_gobp_path   <- "C:/Users/Maggie Drelichman/Thesis/data/gsea/PLACENTA_MALE_GOBP"
# female_gobp_path <- "C:/Users/Maggie Drelichman/Thesis/data/gsea/PLACENTA_FEMALE_GOBP"

# --- KEGG (add when ready) ---
# male_kegg_path   <- "C:/Users/Maggie Drelichman/Thesis/data/gsea/PLACENTA_MALE_KEGG"
# female_kegg_path <- "C:/Users/Maggie Drelichman/Thesis/data/gsea/PLACENTA_FEMALE_KEGG"
```

```{r}
# ============================================================
# HELPER FUNCTION — auto-detects TSV filename (no hardcoded timestamps)
# ============================================================

read_gsea_tsv <- function(folder, direction, sex, database) {
  pattern <- if (direction == "pos") {
    "gsea_report_for_na_pos.*\\.tsv"
  } else {
    "gsea_report_for_na_neg.*\\.tsv"
  }

  f <- list.files(folder, pattern = pattern, full.names = TRUE)[1]

  if (is.na(f)) {
    warning("No TSV found in: ", folder, " for direction: ", direction)
    return(NULL)
  }

  read.delim(f, stringsAsFactors = FALSE) %>%
    mutate(
      sex       = sex,
      database  = database,
      direction = if (direction == "pos") "Upregulated" else "Downregulated"
    )
}
```

```{r}
# ============================================================
# STEP 2 — Read all GSEA results
# ============================================================

male_h_pos   <- read_gsea_tsv(male_hallmark_path,   "pos", "Male",   "Hallmark")
male_h_neg   <- read_gsea_tsv(male_hallmark_path,   "neg", "Male",   "Hallmark")
female_h_pos <- read_gsea_tsv(female_hallmark_path, "pos", "Female", "Hallmark")
female_h_neg <- read_gsea_tsv(female_hallmark_path, "neg", "Female", "Hallmark")

# Uncomment when additional collections are ready:
# male_r_pos   <- read_gsea_tsv(male_reactome_path,   "pos", "Male",   "Reactome")
# male_r_neg   <- read_gsea_tsv(male_reactome_path,   "neg", "Male",   "Reactome")
# female_r_pos <- read_gsea_tsv(female_reactome_path, "pos", "Female", "Reactome")
# female_r_neg <- read_gsea_tsv(female_reactome_path, "neg", "Female", "Reactome")

# Bind all collections together
all_gsea <- bind_rows(
  male_h_pos, male_h_neg,
  female_h_pos, female_h_neg
  # male_r_pos, male_r_neg,
  # female_r_pos, female_r_neg
)

colnames(all_gsea)
cat("Total rows loaded:", nrow(all_gsea), "\n")
```

```{r}
# ============================================================
# STEP 3 — Clean pathway names & filter significant pathways
# Broad threshold: FDR < 0.25 and NOM p < 0.05
# ============================================================

all_gsea_clean <- all_gsea %>%
  filter(!is.na(NAME), NAME != "") %>%
  mutate(
    pathway_label = NAME %>%
      str_remove("^HALLMARK_") %>%
      str_remove("^REACTOME_") %>%
      str_remove("^GOBP_") %>%
      str_remove("^KEGG_") %>%
      str_replace_all("_", " ") %>%
      str_to_title() %>%
      str_wrap(width = 40),
    sex = factor(sex, levels = c("Female", "Male"))
  )

sig_gsea <- all_gsea_clean %>%
  filter(FDR.q.val < 0.25, NOM.p.val < 0.05)

cat("Significant pathways total:", nrow(sig_gsea), "\n")
print(table(sig_gsea$sex, sig_gsea$direction, sig_gsea$database))
```

# Visualization 1: Bubble / Dot Plot (Sex Comparison)
Pathways on y-axis, sex on x-axis. Color = NES (blue = up, red = down). Size = -log10(FDR). Only pathways significant in at least one sex are shown.

```{r fig.width=9, fig.height=10}
# ============================================================
# All pathways significant in at least one sex
# ============================================================

# Get union of significant pathway names per database
sig_names <- sig_gsea %>%
  select(NAME, database) %>%
  distinct()

# Pull full data for those pathways across both sexes
bubble_data <- all_gsea_clean %>%
  inner_join(sig_names, by = c("NAME", "database")) %>%
  mutate(
    sig_label = case_when(
      FDR.q.val < 0.05  ~ "FDR < 0.05",
      FDR.q.val < 0.25  ~ "FDR < 0.25",
      TRUE               ~ "ns"
    ),
    sig_label = factor(sig_label,
                       levels = c("FDR < 0.05", "FDR < 0.25", "ns"))
  )

bubble_plot <- ggplot(
  bubble_data,
  aes(x     = sex,
      y     = fct_reorder(pathway_label, NES),
      color = NES,
      size  = -log10(FDR.q.val + 1e-4),
      shape = sig_label)
) +
  geom_point(alpha = 0.85) +
  scale_color_gradient2(
    low      = "#C0392B",
    mid      = "white",
    high     = "#2980B9",
    midpoint = 0,
    name     = "NES"
  ) +
  scale_size_continuous(
    name   = "-Log10(FDR)",
    range  = c(2, 8)
  ) +
  scale_shape_manual(
    values = c("FDR < 0.05" = 16, "FDR < 0.25" = 17, "ns" = 1),
    name   = "Significance"
  ) +
  facet_wrap(~ database, scales = "free_y", ncol = 1) +
  labs(
    title = "GSEA Pathway Enrichment by Sex\n20uM PFBS vs. Control — Placenta",
    x     = NULL,
    y     = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold", hjust = 0.5, size = 12),
    axis.text.y  = element_text(size = 8, color = "black"),
    axis.text.x  = element_text(size = 11, face = "bold"),
    strip.text   = element_text(face = "bold"),
    panel.grid.major.x = element_blank()
  )

print(bubble_plot)

ggsave(
  filename    = "Figure_GSEA_BubblePlot_Sex.tiff",
  plot        = bubble_plot,
  path        = "C:/Users/Maggie Drelichman/Thesis/output/figures",
  width       = 9,
  height      = 10,
  units       = "in",
  dpi         = 1200,
  device      = "tiff",
  compression = "lzw"
)
```

# Visualization 2: Grouped NES Bar Chart (Sex Comparison)
Side-by-side bars per pathway. Only significant pathways (FDR < 0.25) shown. Colored by sex.

```{r fig.width=10, fig.height=8}
# ============================================================
# Use only pathways significant in at least one sex
# ============================================================

bar_data <- bubble_data %>%
  mutate(NES_plot = ifelse(is.na(NES), 0, NES))

bar_plot <- ggplot(
  bar_data,
  aes(x    = NES_plot,
      y    = fct_reorder(pathway_label, NES_plot),
      fill = sex)
) +
  geom_col(
    position = position_dodge(width = 0.7),
    width    = 0.65,
    alpha    = 0.85
  ) +
  geom_vline(xintercept = 0, color = "grey40", linewidth = 0.5) +
  scale_fill_manual(
    values = c("Female" = "#E07BA0", "Male" = "#5B8DB8"),
    name   = "Sex"
  ) +
  facet_wrap(~ database, scales = "free_y", ncol = 1) +
  labs(
    title = "GSEA NES by Sex — Placenta 20uM PFBS",
    x     = "Normalized Enrichment Score (NES)",
    y     = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", hjust = 0.5, size = 12),
    axis.text.y        = element_text(size = 8, color = "black"),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.4),
    legend.position    = "bottom",
    strip.text         = element_text(face = "bold")
  )

print(bar_plot)

ggsave(
  filename    = "Figure_GSEA_BarChart_Sex.tiff",
  plot        = bar_plot,
  path        = "C:/Users/Maggie Drelichman/Thesis/output/figures",
  width       = 10,
  height      = 8,
  units       = "in",
  dpi         = 1200,
  device      = "tiff",
  compression = "lzw"
)
```

# Visualization 3: Venn Diagram
Overlap of significant pathway names between male and female. Good for a quick 2-set comparison. Use per collection or combined.

```{r fig.width=7, fig.height=6}
# ============================================================
# Significant pathway names per sex
# ============================================================

male_sig_pathways   <- sig_gsea %>% filter(sex == "Male")   %>% pull(NAME) %>% unique()
female_sig_pathways <- sig_gsea %>% filter(sex == "Female") %>% pull(NAME) %>% unique()

venn_list <- list(
  Male   = male_sig_pathways,
  Female = female_sig_pathways
)

cat("Male significant pathways:  ", length(male_sig_pathways), "\n")
cat("Female significant pathways:", length(female_sig_pathways), "\n")
cat("Shared pathways:", length(intersect(male_sig_pathways, female_sig_pathways)), "\n")
cat("\nShared pathway names:\n")
print(intersect(male_sig_pathways, female_sig_pathways))

venn_plot <- ggVennDiagram(
  venn_list,
  label_alpha = 0,
  set_color   = c("#5B8DB8", "#E07BA0")
) +
  scale_fill_gradient(low = "white", high = "#B0C4DE") +
  labs(title = "Shared Significant Pathways: Male vs. Female Placenta\n20uM PFBS — FDR < 0.25") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11))

print(venn_plot)

ggsave(
  filename    = "Figure_GSEA_Venn_Sex.tiff",
  plot        = venn_plot,
  path        = "C:/Users/Maggie Drelichman/Thesis/output/figures",
  width       = 7,
  height      = 6,
  units       = "in",
  dpi         = 1200,
  device      = "tiff",
  compression = "lzw"
)
```

# Visualization 4: UpSet Plot
Better than Venn when you have multiple collections. Shows all combinations of overlap. Extend the list below as you add more collections.

```{r fig.width=9, fig.height=6}
# ============================================================
# Build membership matrix across groups
# Add more groups here as you run more collections
# ============================================================

all_pathways <- unique(sig_gsea$NAME)

upset_data <- data.frame(
  pathway = all_pathways,
  Male    = as.integer(all_pathways %in% male_sig_pathways),
  Female  = as.integer(all_pathways %in% female_sig_pathways)

  # Add more columns when ready, e.g.:
  # Male_Reactome   = as.integer(all_pathways %in% male_reactome_sig),
  # Female_Reactome = as.integer(all_pathways %in% female_reactome_sig)
)

# Save a copy — useful to inspect which pathways are in which intersections
write.csv(
  upset_data,
  "C:/Users/Maggie Drelichman/Thesis/output/placenta/GSEA_pathway_overlap_matrix.csv",
  row.names = FALSE
)

tiff(
  filename    = "C:/Users/Maggie Drelichman/Thesis/output/figures/Figure_GSEA_UpSet_Sex.tiff",
  width       = 9,
  height      = 6,
  units       = "in",
  res         = 300,
  compression = "lzw"
)

upset(
  upset_data,
  sets            = c("Male", "Female"),
  order.by        = "freq",
  sets.bar.color  = c("#5B8DB8", "#E07BA0"),
  main.bar.color  = "grey30",
  text.scale      = 1.3,
  mainbar.y.label = "Pathway Intersections",
  sets.x.label    = "Significant Pathways per Sex"
)

dev.off()

# Also print to viewer
upset(
  upset_data,
  sets            = c("Male", "Female"),
  order.by        = "freq",
  sets.bar.color  = c("#5B8DB8", "#E07BA0"),
  main.bar.color  = "grey30",
  text.scale      = 1.3,
  mainbar.y.label = "Pathway Intersections",
  sets.x.label    = "Significant Pathways per Sex"
)
```

# Visualization 5: EnrichmentMap (Cytoscape — export files here)
EnrichmentMap is built in Cytoscape, not R. Run this chunk to export the files GSEA already created in the correct format. Then in Cytoscape: Apps > EnrichmentMap > load the files below.

```{r}
# ============================================================
# GSEA already produces the files EnrichmentMap needs.
# This chunk just reminds you what to load in Cytoscape.
# ============================================================

cat("=== FILES TO LOAD IN CYTOSCAPE EnrichmentMap ===\n\n")

cat("For MALE Hallmark:\n")
cat("  Results file (pos): ", list.files(male_hallmark_path, "gsea_report_for_na_pos.*tsv", full.names=TRUE)[1], "\n")
cat("  Results file (neg): ", list.files(male_hallmark_path, "gsea_report_for_na_neg.*tsv", full.names=TRUE)[1], "\n")
cat("  GMT file: the h.all.*.symbols.gmt you used for GSEA\n\n")

cat("For FEMALE Hallmark:\n")
cat("  Results file (pos): ", list.files(female_hallmark_path, "gsea_report_for_na_pos.*tsv", full.names=TRUE)[1], "\n")
cat("  Results file (neg): ", list.files(female_hallmark_path, "gsea_report_for_na_neg.*tsv", full.names=TRUE)[1], "\n")
cat("  GMT file: the h.all.*.symbols.gmt you used for GSEA\n\n")

cat("In EnrichmentMap: load male as Dataset 1, female as Dataset 2.\n")
cat("Set FDR cutoff to 0.25, overlap cutoff to 0.375 (Jaccard+Overlap combined).\n")
cat("Nodes colored by NES, edges by gene overlap.\n")
```

```{r}
# ============================================================
# EXPORT: Save significant pathway table for reference
# ============================================================

write.csv(
  sig_gsea %>%
    arrange(database, sex, direction, desc(abs(NES))) %>%
    select(NAME, sex, direction, NES, NOM.p.val, FDR.q.val, SIZE, LEADING.EDGE, database),
  file      = "C:/Users/Maggie Drelichman/Thesis/output/placenta/GSEA_placenta_significant_pathways.csv",
  row.names = FALSE
)

cat("Significant pathway table saved!\n")
cat("Total significant pathways:", nrow(sig_gsea), "\n")
print(table(sig_gsea$sex, sig_gsea$direction))
```
