---
title: "Term_2026_20"
output: html_document
date: "2026-04-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Load Packages 
```{r}
library( "DESeq2" )
library(here)
library(tidyverse)
library(ggplot2)
```
# Load Feature Count Data 
```{r}
# Define the file path and set working directory 
setwd(here("data", "featurecounts_2026"))

# n1_20um_26 
n1_20um_26_count_data <- read.csv('n1_20um_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n1_20um_26_count_data)[2] = "n1_20um_26"

# n1_control_26 

n1_control_26_count_data <- read.csv('n1_control_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n1_control_26_count_data)[2] = "n1_control_26"

# n2_20um_26 
n2_20um_26_count_data <- read.csv('n2_20um_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n2_20um_26_count_data)[2] = "n2_20um_26"

# n2_control_26 

n2_control_26_count_data <- read.csv('n2_control_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n2_control_26_count_data)[2] = "n2_control_26"

# n3_20um_26 

n3_20um_26_count_data <- read.csv('n3_20um_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n3_20um_26_count_data)[2] = "n3_20um_26"

# n3_control_26 

n3_control_26_count_data <- read.csv('n3_control_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n3_control_26_count_data)[2] = "n3_control_26"


# n4_20um_26 

n4_20um_26_count_data <- read.csv('n4_20um_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n4_20um_26_count_data)[2] = "n4_20um_26"

# n4_control_26 

n4_control_26_count_data <- read.csv('n4_control_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n4_control_26_count_data)[2] = "n4_control_26"

# n5_20um_26 

n5_20um_26_count_data <- read.csv('n5_20um_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n5_20um_26_count_data)[2] = "n5_20um_26"

# n5_control_26 

n5_control_26_count_data <- read.csv('n5_control_26 featureCounts.tabular', header = TRUE, sep = "\t")
colnames(n5_control_26_count_data)[2] = "n5_control_26"

```

# Join all count data into one data frame by full_join & change all "NA values" to 0 
```{r}
# put all data frames into list 

df_list <- list(n1_20um_26_count_data, n1_control_26_count_data, n2_20um_26_count_data, n2_control_26_count_data, n3_20um_26_count_data, n3_control_26_count_data, n4_20um_26_count_data, n4_control_26_count_data, n5_20um_26_count_data, n5_control_26_count_data)

library(purrr)

# merge all data frames into list 

count_data <- purrr::reduce(df_list, dplyr::full_join, by= 'Geneid')


# make the Geneid the row id 
count_data <- count_data %>% remove_rownames %>% column_to_rownames(var="Geneid")
count_data[is.na(count_data)] <- 0
head(count_data)

# Write Count Data as .csv to File Path 
count_20um_26_file_path <- here("output", "placenta", "count_20um_26.csv")

write.csv(as.data.frame(count_data), file = count_20um_26_file_path, row.names = TRUE)
```
# Create Meta Data Table 
# Ensuring the rows in meta_data match the columns in counts_data 
## Goal: Comparing "control" vs. "treated" across five donors (trials; N=5)

```{r}

SampleName <- c(colnames(count_data))
treatment <- c("treated","control","treated","control","treated","control","treated","control","treated","control")
trial <- c("one","one","two","two","three","three","four","four","five","five")
sex <-c("F","F","F","F", "M","M","M","M","F","F")

meta_data <- data.frame(SampleName, treatment, trial, sex)
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="SampleName")
meta_data$treatment <- as.factor(meta_data$treatment)
meta_data$trial <- as.factor(meta_data$trial)
meta_data$sex <- as.factor(meta_data$sex)


all(colnames(count_data) %in% rownames(meta_data))
all(colnames(count_data) == rownames(meta_data))
```
# Setting Up DESeq2 Objects 
# Model gene expression as a function treatment while controling for trial differences, variable of interest should always be last in DESeq2 design formulas  
```{r}
design_formula <- ~ trial + treatment 
dds <- DESeqDataSetFromMatrix(
  countData = count_data,
  colData   = meta_data,
  design    = design_formula
)
# shows object elements
dim(dds)
```
# Investigate DESeq Object
```{r}
dds
```
# Pre-filtering: by removing rows in which there are very few reads
# Filter any counts less than 200 
```{r}
keep <- rowSums(counts(dds)) >= 200
dds <- dds[keep,]
dds
head(dds)
```
# Set Reference Factor; ensure we are comparing against the control 
```{r}

# Setting ref = 'one' defines n1 as the intercept in the linear model.
# DESeq2 estimates donor-specific offsets for trials two through five, effectively blocking the donor effect so that the treatment comparison is made within the donors.
# This is so the treatment comparison is made correctly within each donor rather than across them 

dds$treatment <- relevel(dds$treatment, ref = "control")
dds$trial     <- relevel(dds$trial,     ref = "one")


design(dds)
```

# Create PCA Plots 

# PCA plot

```{r}
library(DESeq2)
library(ggplot2)

# Transform using rlog
rld.all <- rlog(dds, blind = TRUE)

# Relabel factors BEFORE plotting
colData(rld.all)$treatment <- factor(
  colData(rld.all)$treatment,
  levels = c("control", "treated"),
  labels = c("Control", "20 µM")
)

colData(rld.all)$trial <- factor(
  colData(rld.all)$trial,
  levels = c("one", "two", "three", "four", "five"),
  labels = c("Placenta 1", "Placenta 2", "Placenta 3", "Placenta 4", "Placenta 5")
)
# Verify metadata looks correct BEFORE plotting 
head(colData(rld.all))

# Extract PCA data for ggplot 
pcaData <- plotPCA(
  rld.all,
  intgroup = c("treatment", "trial"),
  returnData = TRUE
)
# Extract percent variance 
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Check pcaData looks right
head(pcaData)

# Build Plots 
p1 <- ggplot(pcaData, aes(PC1, PC2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Control" = "#1b9e77", "20 µM" = "#d95f02")) +
  labs(
    title = "Term Placental Tissue: Control vs. Treatment",
    x = paste0("PC1: ", percentVar[1], "% variance"),
    y = paste0("PC2: ", percentVar[2], "% variance"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

p2 <- ggplot(pcaData, aes(PC1, PC2, color = trial)) +
  geom_point(size = 4) +
  scale_color_manual(values = c(
    "Placenta 1" = "#66c2a5",
    "Placenta 2" = "#fc8d62",
    "Placenta 3" = "#8da0cb",
    "Placenta 4" = "#FFF891",
    "Placenta 5" = "#FFBDF7"
  )) +
  labs(
    title = "Term Placental Tissue: By Individual Donor",
    x = paste0("PC1: ", percentVar[1], "% variance"),
    y = paste0("PC2: ", percentVar[2], "% variance"),
    color = "Donor"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
# Save as PDF
pdf(file = here("output", "figures", "2026_20um_pca.pdf"),
    width = 7, height = 6)
print(p1)
print(p2)
dev.off()

# Also print to screen to check
print(p1)
print(p2)
```
```{r}
#Combined Plot 
p_combined <- ggplot(pcaData, aes(PC1, PC2, color = treatment, shape = trial)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Control" = "#1b9e77", "20 µM" = "#d95f02")) +
  scale_shape_manual(values = c(
    "Placenta 1" = 16,
    "Placenta 2" = 17,
    "Placenta 3" = 15,
    "Placenta 4" = 18,
    "Placenta 5" = 8
  )) +
  labs(
    title = "Term Placental Tissue: Control vs. Treatment by Donor",
    x = paste0("PC1: ", percentVar[1], "% variance"),
    y = paste0("PC2: ", percentVar[2], "% variance"),
    color = "Treatment",
    shape = "Donor"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

print(p_combined)
```
# Because data is showing an outlier, now include sex differences in term placental tissue 
```{r}
# Placental gene expression is sexually dimorphic, current PCA shows an outlier, and likely that sex confounds treatment effects. 
```
# Confirm Metadata is intact 
```{r}
# This should show all 10 samples with correct labels
as.data.frame(colData(rld.all))[, c("treatment", "trial", "sex")]
```
```{r}
colData(dds)
colnames(colData(dds))
```
# Run Differential Expression 
```{r}
dds_male <- dds[ , dds$sex == "M"] #subsetting DESeq2 object by sex
dds$treatment <- relevel(dds$treatment, ref = 'control')
dds$trial <- relevel(dds$trial, ref = 'one')

dds <- DESeq(dds) #testing

res.all <- results(dds, independentFiltering = FALSE)  # Extract results table

mcols(res.all)
summary(res.all,alpha=0.05) #viewing summary of DE genes

res.all$gene <- rownames(res.all) #adding actual column with gene names
res.all <- res.all [,c('gene','log2FoldChange','baseMean','lfcSE', 'stat','pvalue','padj')] #reordering columns
rownames(res.all)<-NULL #removing non-column row names (genes)
dim(res.all) #checking dataframe
head(res.all)#checking dataframe

log2foldchange_third_5um_results_file_path <- here("output", "placenta", "differential_expression_20um_26.csv")
write.csv(as.data.frame(res.all), file = log2foldchange_third_5um_results_file_path , row.names = TRUE)

```