---
title: "Term_Placenta_Combined_5"
output: html_document
date: "2026-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Load Packages 
```{r}
#library( "DESeq2" )
library(tidyverse)
library(here)
library(ggplot2)
```
# Load Feature Count Data

```{r}
# Define the file path and set working directory 

setwd(here("data", "placenta"))

# n1_5um_25 
n1_5um_25_count_data <- read.csv('tissue_2025_donor1_5um.tabular', header = TRUE, sep = "\t")
colnames(n1_5um_25_count_data)[2] = "n1_5um_25"

# n1_control_25 

n1_control_25_count_data <- read.csv('tissue_2025_donor1_control.tabular', header = TRUE, sep = "\t")
colnames(n1_control_25_count_data)[2] = "n1_control_25"

# n2_5um_25 
n2_5um_25_count_data <- read.csv('tissue_2025_donor2_5um.tabular', header = TRUE, sep = "\t")
colnames(n2_5um_25_count_data)[2] = "n2_5um_25"

# n2_control_25 

n2_control_25_count_data <- read.csv('tissue_2025_donor2_control.tabular', header = TRUE, sep = "\t")
colnames(n2_control_25_count_data)[2] = "n2_control_25"

# n3_5um_25 
n3_5um_25_count_data <- read.csv('tissue_2025_donor3_5um.tabular', header = TRUE, sep = "\t")
colnames(n3_5um_25_count_data)[2] = "n3_5um_25"

# n3_control_25 

n3_control_25_count_data <- read.csv('tissue_2025_donor3_control.tabular', header = TRUE, sep = "\t")
colnames(n3_control_25_count_data)[2] = "n3_control_25"


# n1_5um_26 
n1_5um_26_count_data <- read.csv('tissue_2026_donor1_5um.tabular', header = TRUE, sep = "\t")
colnames(n1_5um_26_count_data)[2] = "n1_5um_26"

# n1_control_26 

n1_control_26_count_data <- read.csv('tissue_2026_donor1_control.tabular', header = TRUE, sep = "\t")
colnames(n1_control_26_count_data)[2] = "n1_control_26"

# n2_5um_26 
n2_5um_26_count_data <- read.csv('tissue_2026_donor2_5um.tabular', header = TRUE, sep = "\t")
colnames(n2_5um_26_count_data)[2] = "n2_5um_26"

# n2_control_26 

n2_control_26_count_data <- read.csv('tissue_2026_donor2_control.tabular', header = TRUE, sep = "\t")
colnames(n2_control_26_count_data)[2] = "n2_control_26"

# n3_5um_26 

n3_5um_26_count_data <- read.csv('tissue_2026_donor3_5um.tabular', header = TRUE, sep = "\t")
colnames(n3_5um_26_count_data)[2] = "n3_5um_26"

# n3_control_26 

n3_control_26_count_data <- read.csv('tissue_2026_donor3_control.tabular', header = TRUE, sep = "\t")
colnames(n3_control_26_count_data)[2] = "n3_control_26"

# n5_5um_26 

n5_5um_26_count_data <- read.csv('tissue_2026_donor5_5um.tabular', header = TRUE, sep = "\t")
colnames(n5_5um_26_count_data)[2] = "n5_5um_26"

# n5_control_26 

n5_control_26_count_data <- read.csv('tissue_2026_donor5_control.tabular', header = TRUE, sep = "\t")
colnames(n5_control_26_count_data)[2] = "n5_control_26"

```

# Join all count data into one data frame by full_join & change all "NA values" to 0 
```{r}
# put all data frames into list 

df_list <- list(n1_5um_25_count_data, n1_control_25_count_data, n2_5um_25_count_data, n2_control_25_count_data, n3_5um_25_count_data, n3_control_25_count_data, n1_5um_26_count_data, n1_control_26_count_data, n2_5um_26_count_data, n2_control_26_count_data, n3_5um_26_count_data, n3_control_26_count_data, n5_5um_26_count_data, n5_control_26_count_data)

library(purrr)

# merge all data frames into list 

count_data <- purrr::reduce(df_list, dplyr::full_join, by= 'Geneid')


# make the Geneid the row id 
count_data <- count_data %>% remove_rownames %>% column_to_rownames(var="Geneid")
count_data[is.na(count_data)] <- 0
head(count_data)

# Write Count Data as .csv to File Path 
count_5um_26_file_path <- here("output", "placenta", "combined_count_5um.csv")

write.csv(as.data.frame(count_data), file = count_5um_26_file_path, row.names = TRUE)
```
# Create Meta Data Table 
# Ensuring the rows in meta_data match the columns in counts_data 
## Goal: Comparing "control" vs. "treated" across seven donors (sample_id; N=7)

```{r}

SampleName <- c(colnames(count_data))
treatment <- c("treated","control","treated","control","treated","control","treated","control","treated","control","treated","control","treated","control")
sample_id <- c("one","one","two","two","three","three","four","four","five","five","six","six","eight","eight")
sex <-c("M","M","M","M","M","M","F","F","F","F", "M","M","F","F")
study <-c("first","first","first","first","first","first","second","second","second","second","second","second","second","second")

meta_data <- data.frame(SampleName, treatment, sample_id, sex, study)
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="SampleName")
meta_data$treatment <- as.factor(meta_data$treatment)
meta_data$sample_id <- as.factor(meta_data$sample_id)
meta_data$sex <- as.factor(meta_data$sex)
meta_data$study <- as.factor(meta_data$study)

# Verify counts columns match metadata rows (full dataset)
all(colnames(count_data) %in% rownames(meta_data))
all(colnames(count_data) == rownames(meta_data))
```

# Descriptive Statistics / QC on the Full Cohort (run BEFORE subsetting by sex)
# PCA + sample-distance heatmap across all samples; these justify splitting by sex.
```{r}
library(DESeq2)
library(ggplot2)
library(pheatmap)

# Build the full-cohort DESeqDataSet (this `dds` is also used by the later
# full-cohort diagnostics chunks that reference `dds`).
dds <- DESeqDataSetFromMatrix(
  countData = count_data,
  colData   = meta_data,
  design    = ~ treatment
)
dds <- estimateSizeFactors(dds)   # populates sizeFactors(dds) for downstream QC

# Normalize for visualization only
vsd <- vst(dds, blind = TRUE)

# PCA — color by every potential confounder
plotPCA(vsd, intgroup = "treatment")
plotPCA(vsd, intgroup = "study")        # batch/sequencing date
plotPCA(vsd, intgroup = "sex")
plotPCA(vsd, intgroup = "sample_id")    # individual donor

# Sample distance heatmap
sampleDists <- dist(t(assay(vsd)))
pheatmap(as.matrix(sampleDists),
         annotation_col = meta_data[, c("treatment","study","sex","sample_id")])
```

```{r}
# Subset metadata by sex
meta_male <- meta_data[meta_data$sex == "M", ]
meta_female <- meta_data[meta_data$sex == "F", ]

# Subset count data to matching samples (keep only columns for that sex)
count_male <- count_data[, rownames(meta_male)]
count_female <- count_data[, rownames(meta_female)]

# Verify alignment
all(colnames(count_male) == rownames(meta_male))
all(colnames(count_female) == rownames(meta_female))
```
```{r}
# Check pairing of treatment
table(meta_male$sample_id, meta_male$treatment)

table(meta_female$sample_id, meta_female$treatment)

# See full metadata
print(meta_male)
print(meta_female)
```
`
# Setting Up DESeq2 Objects 
# Model gene expression as a function treatment while controling for trial differences, variable of interest should always be last in DESeq2 design formulas  
```{r}
meta_male   <- droplevels(meta_male)
meta_female <- droplevels(meta_female)

design_formula_male <- ~ sample_id + treatment 

dds_male <- DESeqDataSetFromMatrix(
  countData = count_male,
  colData   = meta_male,
  design    = design_formula_male
)

design_formula_female <- ~ sample_id + treatment 

dds_female <- DESeqDataSetFromMatrix(
  countData = count_female,
  colData   = meta_female,
  design    = design_formula_female
)
# shows object elements
# dim(dds)


```

# Investigate DESeq Object
```{r}
table(meta_female$study)
table(meta_female$sample_id)
```
```{r}
dds_male
dds_female
```
# Pre-filtering: by removing rows in which there are very few reads
# Filter any counts less than 10

```{r}
keep <- rowSums(counts(dds_male)) >= 200
dds_male <- dds_male[keep,]
dds_male

dds_male <- estimateSizeFactors(dds_male)

keep <- rowSums(counts(dds_female)) >= 200
dds_female <- dds_female[keep,]
dds_female

dds_female <- estimateSizeFactors(dds_female)
```

<!-- The "Diagnose Source of Background Noise & Visualize Variance Structure" step
     was moved earlier: it now runs on the full cohort, before subsetting by sex. -->



# Set Reference Factor; ensure we are comparing against the control 
```{r}

# Setting ref = 'one' defines n1 as the intercept in the linear model.
# DESeq2 estimates donor-specific offsets for trials two through five, effectively blocking the donor effect so that the treatment comparison is made within the donors.
# This is so the treatment comparison is made correctly within each donor rather than across them 

dds_male$treatment <- relevel(dds_male$treatment, ref = "control")

dds_female$treatment <- relevel(dds_female$treatment, ref = "control")

design(dds_male)
design(dds_female)
```

# Create PCA Plots 

# PCA plot

```{r}
library(DESeq2)
library(ggplot2)

# Transform using rlog
rld.all <- rlog(dds, blind = TRUE)

# Relabel factors BEFORE plotting
colData(rld.all)$treatment <- factor(
  colData(rld.all)$treatment,
  levels = c("control", "treated"),
  labels = c("Control", "20 µM")
)

colData(rld.all)$trial <- factor(
  colData(rld.all)$trial,
  levels = c("six","seven","eight","one", "two", "three", "four", "five"),
  labels = c("Placenta 6", "Placenta 7", "Placenta 8","Placenta 1", "Placenta 2", "Placenta 3", "Placenta 4", "Placenta 5")
)
# Verify metadata looks correct BEFORE plotting 
head(colData(rld.all))

# Extract PCA data for ggplot 
pcaData <- plotPCA(
  rld.all,
  intgroup = c("treatment", "trial"),
  returnData = TRUE
)
# Extract percent variance 
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Check pcaData looks right
head(pcaData)

# Build Plots 
p1 <- ggplot(pcaData, aes(PC1, PC2, color = treatment)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Control" = "#1b9e77", "20 µM" = "#d95f02")) +
  labs(
    title = "Term Placental Tissue: Control vs. Treatment",
    x = paste0("PC1: ", percentVar[1], "% variance"),
    y = paste0("PC2: ", percentVar[2], "% variance"),
    color = "Treatment Group"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

p2 <- ggplot(pcaData, aes(PC1, PC2, color = trial)) +
  geom_point(size = 4) +
  scale_color_manual(values = c(
    "Placenta 1" = "#66c2a5",
    "Placenta 2" = "#fc8d62",
    "Placenta 3" = "#8da0cb",
    "Placenta 4" = "#FFF891",
    "Placenta 5" = "#FFBDF7",
    "Placenta 6" = "#8085ED",
    "Placenta 7" = "#80DBED",
    "Placenta 8" = "#ED80A1"
  )) +
  labs(
    title = "Term Placental Tissue: By Individual Donor",
    x = paste0("PC1: ", percentVar[1], "% variance"),
    y = paste0("PC2: ", percentVar[2], "% variance"),
    color = "Donor"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
# Save as PDF
pdf(file = here("output", "figures", "Combined_20um_pca.pdf"),
    width = 7, height = 6)
print(p1)
print(p2)
dev.off()

# Also print to screen to check
print(p1)
print(p2)
```
```{r}
#Combined Plot 
p_combined <- ggplot(pcaData, aes(PC1, PC2, color = treatment, shape = trial)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Control" = "#1b9e77", "20 µM" = "#d95f02")) +
  scale_shape_manual(values = c(
    "Placenta 1" = 16,
    "Placenta 2" = 17,
    "Placenta 3" = 15,
    "Placenta 4" = 18,
    "Placenta 5" = 8,
    "Placenta 6" = 3,
    "Placenta 7" = 0,
    "Placenta 8" = 20
  )) +
  labs(
    title = "Term Placental Tissue: Control vs. Treatment by Donor",
    x = paste0("PC1: ", percentVar[1], "% variance"),
    y = paste0("PC2: ", percentVar[2], "% variance"),
    color = "Treatment",
    shape = "Donor"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

print(p_combined)
```

# Run Differential Expression 
```{r}
dds_male <- DESeq(dds_male) #testing

res.male <- results(dds_male, independentFiltering = FALSE)  # Extract results table

mcols(res.male)
summary(res.male,alpha=0.05) #viewing summary of DE genes

res.male$gene <- rownames(res.male) #adding actual column with gene names
res.male <- res.male [,c('gene','log2FoldChange','baseMean','lfcSE', 'stat','pvalue','padj')] #reordering columns
rownames(res.male)<-NULL #removing non-column row names (genes)
dim(res.male) #checking dataframe
head(res.male)#checking dataframe

log2foldchange_differential_expression_5um_26_results_file_path <- here("output", "placenta", "male_differential_expression_5um_26.csv")
write.csv(as.data.frame(res.male), file = log2foldchange_differential_expression_5um_26_results_file_path , row.names = TRUE)

```
# Save as res.male as .rnk file for Pathway Analysis in GSEA 
```{r}
res.male_df <- as.data.frame(res.male)
ranked_list <- res.male_df %>%
  filter(!is.na(log2FoldChange)) %>%
  filter(!is.na(gene)) %>%
  filter(gene != "") %>%
  group_by(gene) %>%
  slice_max(
    order_by  = abs(log2FoldChange),
    n         = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  arrange(desc(log2FoldChange)) %>%
  select(gene, log2FoldChange)


head(ranked_list)
tail(ranked_list)
cat("Total genes:", nrow(ranked_list), "\n")


write.table(
  ranked_list,
  file      = here("output", "placenta", "male_5_preranked.rnk"),
  sep       = "\t",          # tab separated
  row.names = FALSE,         # no row numbers
  col.names = FALSE,         # no header
  quote     = FALSE,         # no quotes
  fileEncoding = "UTF-8"     # standard encoding GSEA can read
)

cat("RNK file saved!\n")

# Step 4 - verify file looks correct
# open in Notepad NOT Excel to check
verify <- read.table(
  here("output", "placenta", "male_5_preranked.rnk"),
  sep = "\t",
  stringsAsFactors = FALSE
)
```

```{r}

dds_female <- DESeq(dds_female) #testing

res.female <- results(dds_female, independentFiltering = FALSE)  # Extract results table

mcols(res.female)
summary(res.female,alpha=0.05) #viewing summary of DE genes

res.female$gene <- rownames(res.female) #adding actual column with gene names
res.female <- res.female [,c('gene','log2FoldChange','baseMean','lfcSE', 'stat','pvalue','padj')] #reordering columns
rownames(res.female)<-NULL #removing non-column row names (genes)
dim(res.female) #checking dataframe
head(res.female)#checking dataframe

F_log2foldchange_differential_expression_5um_26_results_file_path <- here("output", "placenta", "female_differential_expression_5um_26.csv")
write.csv(as.data.frame(res.female), file = F_log2foldchange_differential_expression_5um_26_results_file_path , row.names = TRUE)

```
```{r}
res.female_df <- as.data.frame(res.female)
ranked_list <- res.female_df %>%
  filter(!is.na(log2FoldChange)) %>%
  filter(!is.na(gene)) %>%
  filter(gene != "") %>%
  group_by(gene) %>%
  slice_max(
    order_by  = abs(log2FoldChange),
    n         = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  arrange(desc(log2FoldChange)) %>%
  select(gene, log2FoldChange)


head(ranked_list)
tail(ranked_list)
cat("Total genes:", nrow(ranked_list), "\n")


write.table(
  ranked_list,
  file      = here("output", "placenta", "female_5_preranked.rnk"),
  sep       = "\t",          # tab separated
  row.names = FALSE,         # no row numbers
  col.names = FALSE,         # no header
  quote     = FALSE,         # no quotes
  fileEncoding = "UTF-8"     # standard encoding GSEA can read
)

cat("RNK file saved!\n")

# Step 4 - verify file looks correct
# open in Notepad NOT Excel to check
verify <- read.table(
  here("output", "placenta", "female_5_preranked.rnk"),
  sep = "\t",
  stringsAsFactors = FALSE
)
```

# New Threshold Criteria: Getting the # of DEG which meet both criteria 

```{r}
cat("Max -log10(padj):", max(-log10(res.female$padj), na.rm = TRUE), "\n")
cat("Max LFC:", max(res.female$log2FoldChange, na.rm = TRUE), "\n")
cat("Min LFC:", min(res.female$log2FoldChange, na.rm = TRUE), "\n")

cat("Upregulated (padj<0.05, LFC>0.585):", 
    sum(res.female$padj < 0.05 & res.female$log2FoldChange > log2(1.5), na.rm = TRUE), "\n")
cat("Downregulated (padj<0.05, LFC< -0.585):", 
    sum(res.female$padj < 0.05 & res.female$log2FoldChange < -log2(1.5), na.rm = TRUE), "\n")
cat("Significant |LFC|<0.585:", 
    sum(res.female$padj < 0.05 & abs(res.female$log2FoldChange) <= log2(1.5), na.rm = TRUE), "\n")
```


```{r}
hist(res.male$pvalue, breaks = 50,
     main = "P-value Distribution Male",
     xlab = "p-value",
     col = "steelblue")
hist(res.female$pvalue, breaks = 50,
     main = "P-value Distribution Female",
     xlab = "p-value",
     col = "red")
```
Max -log10(padj): 10.80529 
Max LFC: 2.778603 
Min LFC: -2.795013 
Upregulated (padj<0.05, LFC>0.585): 45 
Downregulated (padj<0.05, LFC< -0.585): 14 
Significant |LFC|<0.585: 62 
Volcano Plot 
```{r}

```

```{r}
head(res.male)
res.male_df <- as.data.frame(res.male)
"gene" %in% colnames(res.male_df)
library(ggplot2)
library(ggrepel)
library(dplyr)

# ============================================
# STEP 1 — Prepare volcano data
# ============================================

volcano_data_male <- res.male_df %>%
  filter(!is.na(pvalue)) %>%
  filter(!is.na(padj)) %>%
  filter(!is.na(log2FoldChange)) %>%
  mutate(
    neg_log10_padj = pmin(-log10(padj), 11),

    significance = case_when(
      padj < 0.05 & log2FoldChange >  log2(1.5) ~ "Upregulated",
      padj < 0.05 & log2FoldChange < -log2(1.5) ~ "Downregulated",
      padj < 0.05                                ~ "Significant (|LFC| < 0.585)",
      TRUE                                       ~ "Not Significant"
    ),

    significance = factor(
      significance,
      levels = c(
        "Upregulated",
        "Downregulated",
        "Significant (|LFC| < 0.585)",
        "Not Significant"
      )
    )
  )

cat("Gene counts by category:\n")
print(table(volcano_data_male$significance))

# ============================================
# STEP 2 — Select genes to label
# ============================================

top_up_male <- volcano_data_male %>%
  filter(significance == "Upregulated") %>%
  arrange(padj) %>%
  head(5)

top_down_male <- volcano_data_male %>%
  filter(significance == "Downregulated") %>%
  arrange(padj) %>%
  head(5)

top_sig_male <- volcano_data_male %>%
  filter(significance == "Significant (|LFC| < 0.585)") %>%
  arrange(padj) %>%
  head(10)

if (nrow(top_up_male) > 0 | nrow(top_down_male) > 0) {
  genes_to_label_male <- bind_rows(top_up_male, top_down_male)
  cat("\nLabeling up and downregulated genes\n")
} else {
  genes_to_label_male <- top_sig_male
  cat("\nLabeling top significant genes\n")
}

cat("Genes labeled:", nrow(genes_to_label_male), "\n")
print(genes_to_label_male$gene)

# ============================================
# STEP 3 — Teal gradient palette
# ============================================

teal_colors <- c(
  "Upregulated"                 = "#004c4c",
  "Downregulated"               = "#008080",
  "Significant (|LFC| < 0.585)" = "#5ba8a8",
  "Not Significant"             = "#d4eaea"
)

# ============================================
# STEP 4 — Build plot
# ============================================

volcano_plot_male <- ggplot(
  volcano_data_male,
  aes(x     = log2FoldChange,
      y     = neg_log10_padj,
      color = significance)
) +
  geom_point(
    data  = volcano_data_male %>% filter(significance == "Not Significant"),
    alpha = 0.4,
    size  = 1.5
  ) +
  geom_point(
    data  = volcano_data_male %>% filter(significance != "Not Significant"),
    alpha = 0.9,
    size  = 2.0
  ) +
  scale_color_manual(
    values = teal_colors,
    name   = "Expression",
    labels = c(
      "Upregulated" = paste0(
        "Upregulated (padj<0.05, LFC>0.585)  n=",
        sum(volcano_data_male$significance == "Upregulated", na.rm = TRUE)
      ),
      "Downregulated" = paste0(
        "Downregulated (padj<0.05, LFC<\u22120.585)  n=",
        sum(volcano_data_male$significance == "Downregulated", na.rm = TRUE)
      ),
      "Significant (|LFC| < 0.585)" = paste0(
        "Significant (padj<0.05, |LFC|<0.585)  n=",
        sum(volcano_data_male$significance == "Significant (|LFC| < 0.585)", na.rm = TRUE)
      ),
      "Not Significant" = paste0(
        "Not Significant  n=",
        sum(volcano_data_male$significance == "Not Significant", na.rm = TRUE)
      )
    )
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype   = "dashed",
    color      = "black",
    linewidth  = 0.5
  ) +
  geom_vline(
    xintercept = c(-log2(1.5), log2(1.5)),
    linetype   = "dashed",
    color      = "black",
    linewidth  = 0.5
  ) +
  geom_label_repel(
    data          = genes_to_label_male,
    aes(label     = gene),
    size          = 2.8,
    max.overlaps  = 20,
    box.padding   = 0.5,
    point.padding = 0.3,
    segment.color = "grey40",
    segment.size  = 0.3,
    show.legend   = FALSE,
    color         = "black",
    fill          = "white",
    label.size    = 0.25
  ) +
  labs(
    title    = "Male Term Placenta: Control vs. 5 µM PFBS",
    subtitle = paste0(
      "Upregulated: ",
      sum(volcano_data_male$significance == "Upregulated", na.rm = TRUE),
      "  |  Downregulated: ",
      sum(volcano_data_male$significance == "Downregulated", na.rm = TRUE),
      "  |  Significant (|LFC|<0.585): ",
      sum(volcano_data_male$significance == "Significant (|LFC| < 0.585)", na.rm = TRUE)
    ),
    x = expression(Log[2]~Fold~Change),
    y = expression(-Log[10]~adjusted~italic(p)~value)
  ) +
  scale_x_continuous(
    limits = c(-6, 6),
    breaks = seq(-6, 6, by = 2)
  ) +
  scale_y_continuous(
    limits = c(0, 11),
    breaks = seq(0, 11, by = 2)
  ) +
  theme_bw() +
  theme(
    plot.title       = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle    = element_text(hjust = 0.5, size = 9,  color = "grey40"),
    axis.title.x     = element_text(size = 11),
    axis.title.y     = element_text(size = 11),
    axis.text.x      = element_text(size = 9),
    axis.text.y      = element_text(size = 9),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.text      = element_text(size = 8),
    legend.position  = "bottom",
    legend.direction = "vertical",
    plot.margin      = margin(t = 10, r = 15, b = 10, l = 10, unit = "pt")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 3, alpha = 1))
  )

print(volcano_plot_male)

# ============================================
# STEP 5 — Save plots
# ============================================

ggsave(
  filename    = "New_Threshold_Figure_Volcano_Term_Placenta_5um_PFBS_Male.tiff",
  plot        = volcano_plot_male,
  path        = here("output", "figures"),
  width       = 5,
  height      = 6.5,
  units       = "in",
  dpi         = 1200,
  device      = "tiff",
  compression = "lzw"
)

ggsave(
  filename = "New_Threshold_Figure_Volcano_Term_Placenta_5um_PFBS_Male.pdf",
  plot     = volcano_plot_male,
  path     = here("output", "figures"),
  width    = 5,
  height   = 6.5,
  units    = "in",
  device   = "pdf"
)

cat("\n=== MALE VOLCANO PLOT SAVED ===\n")
cat("Threshold: log2(1.5) = 0.585 (1.5-fold change)\n")
cat("Y axis: 0 to 11 (max is 10.8)\n")
cat("X axis: -6 to 6 (range is -2.8 to 2.8)\n")

# ============================================
# STEP 6 — Export DEG table (both thresholds met)
# padj < 0.05 AND |log2FoldChange| > log2(1.5)
# ============================================

both_thresholds_male_5um <- volcano_data_male %>%
  filter(significance %in% c("Upregulated", "Downregulated")) %>%
  arrange(padj) %>%
  dplyr::select(
    gene,
    log2FoldChange,
    baseMean,
    lfcSE,
    stat,
    pvalue,
    padj,
    significance
  )

cat("\n=== BOTH THRESHOLDS MET — MALE 5µM ===\n")
cat("Upregulated:  ", sum(both_thresholds_male_5um$significance == "Upregulated"),   "\n")
cat("Downregulated:", sum(both_thresholds_male_5um$significance == "Downregulated"), "\n")
cat("Total DEGs:   ", nrow(both_thresholds_male_5um), "\n")
print(both_thresholds_male_5um)

write.csv(
  both_thresholds_male_5um,
  file      = here("output", "both_thresholds_male_5um.csv"),
  row.names = FALSE
)

```

Max -log10(padj): 3.937339 
Max LFC: 1.34622 
Min LFC: -2.031607 
Upregulated (padj<0.05, LFC>0.585): 2 
Downregulated (padj<0.05, LFC< -0.585): 2 
Significant |LFC|<0.585: 0 
```{r}
res.female_df <- as.data.frame(res.female)

library(ggplot2)
library(ggrepel)
library(dplyr)

# ============================================
# STEP 1 — Prepare volcano data
# Max -log10(padj) = 4.58  → y cap at 5
# Max |LFC| = 3.58         → x axis -6 to 6
# ============================================

volcano_data_female <- res.female_df %>%
  filter(!is.na(pvalue)) %>%
  filter(!is.na(padj)) %>%
  filter(!is.na(log2FoldChange)) %>%
  mutate(
    neg_log10_padj = pmin(-log10(padj), 5),

    significance = case_when(
      padj < 0.05 & log2FoldChange >  log2(1.5) ~ "Upregulated",
      padj < 0.05 & log2FoldChange < -log2(1.5) ~ "Downregulated",
      padj < 0.05                                ~ "Significant (|LFC| < 0.585)",
      TRUE                                       ~ "Not Significant"
    ),

    significance = factor(
      significance,
      levels = c(
        "Upregulated",
        "Downregulated",
        "Significant (|LFC| < 0.585)",
        "Not Significant"
      )
    )
  )

cat("Gene counts by category:\n")
print(table(volcano_data_female$significance))

# ============================================
# STEP 2 — Select genes to label
# ============================================

top_up_female <- volcano_data_female %>%
  filter(significance == "Upregulated") %>%
  arrange(padj) %>%
  head(10)

top_down_female <- volcano_data_female %>%
  filter(significance == "Downregulated") %>%
  arrange(padj) %>%
  head(10)

top_sig_female <- volcano_data_female %>%
  filter(significance == "Significant (|LFC| < 0.585)") %>%
  arrange(padj) %>%
  head(10)

if (nrow(top_up_female) > 0 | nrow(top_down_female) > 0) {
  genes_to_label_female <- bind_rows(top_up_female, top_down_female)
  cat("\nLabeling up and downregulated genes\n")
} else {
  genes_to_label_female <- top_sig_female
  cat("\nLabeling top significant genes\n")
}

cat("Genes labeled:", nrow(genes_to_label_female), "\n")
print(genes_to_label_female$gene)

# ============================================
# STEP 3 — Teal gradient palette
# ============================================

teal_colors <- c(
  "Upregulated"                 = "#004c4c",
  "Downregulated"               = "#008080",
  "Significant (|LFC| < 0.585)" = "#5ba8a8",
  "Not Significant"             = "#d4eaea"
)

# ============================================
# STEP 4 — Build plot
# ============================================

volcano_plot_female <- ggplot(
  volcano_data_female,                          # FIX 1
  aes(x     = log2FoldChange,
      y     = neg_log10_padj,
      color = significance)
) +
  geom_point(
    data  = volcano_data_female %>% filter(significance == "Not Significant"),
    alpha = 0.4,
    size  = 1.5
  ) +
  geom_point(
    data  = volcano_data_female %>% filter(significance != "Not Significant"),
    alpha = 0.9,
    size  = 2.0
  ) +
  scale_color_manual(
    values = teal_colors,
    name   = "Expression",
    labels = c(
      "Upregulated" = paste0(
        "Upregulated (padj<0.05, LFC>0.585)  n=",
        sum(volcano_data_female$significance == "Upregulated", na.rm = TRUE)
      ),
      "Downregulated" = paste0(
        "Downregulated (padj<0.05, LFC<\u22120.585)  n=",
        sum(volcano_data_female$significance == "Downregulated", na.rm = TRUE)
      ),
      "Significant (|LFC| < 0.585)" = paste0(
        "Significant (padj<0.05, |LFC|<0.585)  n=",
        sum(volcano_data_female$significance == "Significant (|LFC| < 0.585)", na.rm = TRUE)
      ),
      "Not Significant" = paste0(
        "Not Significant  n=",
        sum(volcano_data_female$significance == "Not Significant", na.rm = TRUE)
      )
    )
  ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype   = "dashed",
    color      = "black",
    linewidth  = 0.5
  ) +
  geom_vline(
    xintercept = c(-log2(1.5), log2(1.5)),
    linetype   = "dashed",
    color      = "black",
    linewidth  = 0.5
  ) +
  geom_label_repel(
    data          = genes_to_label_female,
    aes(label     = gene),
    size          = 2.8,
    max.overlaps  = 20,
    box.padding   = 0.5,
    point.padding = 0.3,
    segment.color = "grey40",
    segment.size  = 0.3,
    show.legend   = FALSE,
    color         = "black",
    fill          = "white",
    label.size    = 0.25
  ) +
  labs(
    title    = "Female Term Placenta: Control vs. 5 µM PFBS",
    subtitle = paste0(
      "Upregulated: ",
      sum(volcano_data_female$significance == "Upregulated", na.rm = TRUE),
      "  |  Downregulated: ",
      sum(volcano_data_female$significance == "Downregulated", na.rm = TRUE),
      "  |  Significant (|LFC|<0.585): ",
      sum(volcano_data_female$significance == "Significant (|LFC| < 0.585)", na.rm = TRUE)
    ),
    x = expression(Log[2]~Fold~Change),
    y = expression(-Log[10]~adjusted~italic(p)~value)
  ) +
  scale_x_continuous(
    limits = c(-6, 6),
    breaks = seq(-6, 6, by = 2)
  ) +
  scale_y_continuous(
    limits = c(0, 5),                           # FIX 3
    breaks = seq(0, 5, by = 1)
  ) +
  theme_bw() +
  theme(
    plot.title       = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle    = element_text(hjust = 0.5, size = 9,  color = "grey40"),
    axis.title.x     = element_text(size = 11),
    axis.title.y     = element_text(size = 11),
    axis.text.x      = element_text(size = 9),
    axis.text.y      = element_text(size = 9),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.text      = element_text(size = 8),
    legend.position  = "bottom",
    legend.direction = "vertical",
    plot.margin      = margin(t = 10, r = 15, b = 10, l = 10, unit = "pt")
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 3, alpha = 1))
  )

print(volcano_plot_female)

# ============================================
# STEP 5 — Save
# ============================================

ggsave(
  filename    = "New_Threshold_Figure_Volcano_Term_Placenta_5um_PFBS_Female.tiff",
  plot        = volcano_plot_female,            # FIX 2a
  path        = here("output", "figures"),
  width       = 5,
  height      = 6.5,
  units       = "in",
  dpi         = 1200,
  device      = "tiff",
  compression = "lzw"
)

ggsave(
  filename = "New_Threshold_Figure_Volcano_Term_Placenta_5um_PFBS_Female.pdf",
  plot     = volcano_plot_female,               # FIX 2b
  path     = here("output", "figures"),
  width    = 5,
  height   = 6.5,
  units    = "in",
  device   = "pdf"
)

cat("\n=== FEMALE VOLCANO PLOT SAVED ===\n")
cat("Threshold: log2(1.5) = 0.585 (1.5-fold change)\n")
cat("Y axis: 0 to 5 (max is 4.58)\n")
cat("X axis: -6 to 6 (range is -3.58 to 1.61)\n")


```{r}
sizeFactors(dds)
```
# Issue in Data Rather than Model- Are there extreme count outlier smaples pulling on dispersion estimates 
```{r}
# Run this now — critical diagnostic
lib_sizes <- colSums(counts(dds))
barplot(lib_sizes,
        las = 2,
        col = ifelse(lib_sizes > mean(lib_sizes) + 2*sd(lib_sizes) |
                     lib_sizes < mean(lib_sizes) - 2*sd(lib_sizes),
                     "red", "steelblue"),
        main = "Library Sizes per Sample",
        ylab = "Total Counts")
abline(h = mean(lib_sizes), lty = 2, col = "red")

# Print exact values
sort(lib_sizes)
```
# Are the treated and control values systamatically different?
```{r}

plotPCA(vst(dds, blind = TRUE), intgroup = "treatment")
```
# What proportion of genes have zero counts?
```{r}
zero_prop <- rowMeans(counts(dds) == 0)
hist(zero_prop, breaks = 50,
     main = "Proportion of Zero Counts per Gene",
     xlab = "Proportion zeros",
     col = "steelblue")

# How many genes are > 50% zeros?
sum(zero_prop > 0.5)
```
#Is the count matrix raw interger counts or have them been transformed? 
```{r}
class(count_data[1,1])   # should be integer or numeric
range(count_data)         # should be 0 to large positive number
head(count_data[, 1:3])  # eyeball the values

```

# Running a Correlation between genes expressed in female & male 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(purrr)

# ============================================
# Load all five results files
# ============================================

male_placenta_5uM   <- read_csv(here("output", "placenta", "male_differential_expression_5um_26.csv"))
male_placenta_20uM  <- read_csv(here("output", "placenta", "male_differential_expression_20um_26.csv"))
female_placenta_5uM  <- read_csv(here("output", "placenta", "female_differential_expression_5um_26.csv"))
female_placenta_20uM <- read_csv(here("output", "placenta", "female_differential_expression_20um_26.csv"))
cells_5um <- read_csv(here("output", "cells", "results_cells_5.csv"))
```

```{r}
# ============================================
# MALE: 5uM vs 20uM dose correlation
# .x = 5uM (first in list), .y = 20uM (second)
# x-axis = 5uM, y-axis = 20uM
# ============================================

df_male_dose <- list(male_placenta_5uM, male_placenta_20uM) %>%
  purrr::reduce(inner_join, by = 'gene')

dim(df_male_dose)
cat("=== MALE: 5uM vs 20uM ===\n")
cor.test(df_male_dose$log2FoldChange.x, df_male_dose$log2FoldChange.y)

# PDF
pdf(file = here("output", "placenta", "male_dose_correlation.pdf"))
smoothScatter(df_male_dose$log2FoldChange.y ~ df_male_dose$log2FoldChange.x,
              colramp = colorRampPalette(c("white", "violet")),
              cex = 3, xlim = c(-5, 5), ylim = c(-5, 5),
              main = "Male Placenta: 5uM vs 20uM PFBS",
              xlab = "Log2Fold Change (5uM)",
              ylab = "Log2Fold Change (20uM)")
abline(v = 0, col = "gray40")
abline(h = 0, col = "gray40")
abline(lm(df_male_dose$log2FoldChange.y ~ df_male_dose$log2FoldChange.x), col = "red", lwd = 2)
dev.off()

# TIFF
tiff(filename = here("output", "placenta", "male_dose_correlation.tiff"),
     width = 6, height = 6, units = "in", res = 300, compression = "lzw")
smoothScatter(df_male_dose$log2FoldChange.y ~ df_male_dose$log2FoldChange.x,
              colramp = colorRampPalette(c("white", "violet")),
              cex = 3, xlim = c(-5, 5), ylim = c(-5, 5),
              main = "Male Placenta: 5uM vs 20uM PFBS",
              xlab = "Log2Fold Change (5uM)",
              ylab = "Log2Fold Change (20uM)")
abline(v = 0, col = "gray40")
abline(h = 0, col = "gray40")
abline(lm(df_male_dose$log2FoldChange.y ~ df_male_dose$log2FoldChange.x), col = "red", lwd = 2)
dev.off()

# ============================================
# FEMALE: 5uM vs 20uM dose correlation
# .x = 5uM (first in list), .y = 20uM (second)
# x-axis = 5uM, y-axis = 20uM
# ============================================

df_female_dose <- list(female_placenta_5uM, female_placenta_20uM) %>%
  purrr::reduce(inner_join, by = 'gene')

dim(df_female_dose)
cat("=== FEMALE: 5uM vs 20uM ===\n")
cor.test(df_female_dose$log2FoldChange.x, df_female_dose$log2FoldChange.y)

# PDF
pdf(file = here("output", "placenta", "female_dose_correlation.pdf"))
smoothScatter(df_female_dose$log2FoldChange.y ~ df_female_dose$log2FoldChange.x,
              colramp = colorRampPalette(c("white", "violet")),
              cex = 3, xlim = c(-5, 5), ylim = c(-5, 5),
              main = "Female Placenta: 5uM vs 20uM PFBS",
              xlab = "Log2Fold Change (5uM)",
              ylab = "Log2Fold Change (20uM)")
abline(v = 0, col = "gray40")
abline(h = 0, col = "gray40")
abline(lm(df_female_dose$log2FoldChange.y ~ df_female_dose$log2FoldChange.x), col = "red", lwd = 2)
dev.off()

# TIFF
tiff(filename = here("output", "placenta", "female_dose_correlation.tiff"),
     width = 6, height = 6, units = "in", res = 300, compression = "lzw")
smoothScatter(df_female_dose$log2FoldChange.y ~ df_female_dose$log2FoldChange.x,
              colramp = colorRampPalette(c("white", "violet")),
              cex = 3, xlim = c(-5, 5), ylim = c(-5, 5),
              main = "Female Placenta: 5uM vs 20uM PFBS",
              xlab = "Log2Fold Change (5uM)",
              ylab = "Log2Fold Change (20uM)")
abline(v = 0, col = "gray40")
abline(h = 0, col = "gray40")
abline(lm(df_female_dose$log2FoldChange.y ~ df_female_dose$log2FoldChange.x), col = "red", lwd = 2)
dev.off()

cat("Both dose correlation plots saved!\n")

```


```{r}
# ============================================
# Cells & Male Term Tissue: 5uM vs 5uM dose correlation
# .x = cells 5uM (first in list), .y = tissue 5uM (second)
# x-axis = cells, y-axis = male tissue
# ============================================

df_cells_male_5dose <- list(cells_5um, male_placenta_5uM) %>%
  purrr::reduce(inner_join, by = 'gene')

dim(df_cells_male_5dose)
cat("=== CELLS vs MALE TISSUE: 5uM ===\n")
cor.test(df_cells_male_5dose$log2FoldChange.x, df_cells_male_5dose$log2FoldChange.y)

# PDF
pdf(file = here("output", "placenta", "cells_male_dose_correlation.pdf"))
smoothScatter(df_cells_male_5dose$log2FoldChange.y ~ df_cells_male_5dose$log2FoldChange.x,
              colramp = colorRampPalette(c("white", "violet")),
              cex = 3, xlim = c(-5, 5), ylim = c(-5, 5),
              main = "5uM PFBS: HTR-8/SVneo vs Male Term Tissue",
              xlab = "Log2Fold Change (Cells 5uM)",
              ylab = "Log2Fold Change (Male Term Tissue 5uM)")
abline(v = 0, col = "gray40")
abline(h = 0, col = "gray40")
abline(lm(df_cells_male_5dose$log2FoldChange.y ~ df_cells_male_5dose$log2FoldChange.x), col = "red", lwd = 2)
dev.off()

# TIFF
tiff(filename = here("output", "placenta", "cells_male_dose_correlation.tiff"),
     width = 6, height = 6, units = "in", res = 300, compression = "lzw")
smoothScatter(df_cells_male_5dose$log2FoldChange.y ~ df_cells_male_5dose$log2FoldChange.x,
              colramp = colorRampPalette(c("white", "violet")),
              cex = 3, xlim = c(-5, 5), ylim = c(-5, 5),
              main = "5uM PFBS: HTR-8/SVneo vs Male Term Tissue",
              xlab = "Log2Fold Change (Cells 5uM)",
              ylab = "Log2Fold Change (Male Term Tissue 5uM)")
abline(v = 0, col = "gray40")
abline(h = 0, col = "gray40")
abline(lm(df_cells_male_5dose$log2FoldChange.y ~ df_cells_male_5dose$log2FoldChange.x), col = "red", lwd = 2)
dev.off()
```

```{r}
# ============================================
# Cells & Female Term Tissue: 5uM vs 5uM dose correlation
# .x = cells 5uM (first in list), .y = tissue 5uM (second)
# x-axis = cells, y-axis = female tissue
# ============================================

df_cells_female_5dose <- list(cells_5um, female_placenta_5uM) %>%
  purrr::reduce(inner_join, by = 'gene')

dim(df_cells_female_5dose)
cat("=== CELLS vs FEMALE TISSUE: 5uM ===\n")
cor.test(df_cells_female_5dose$log2FoldChange.x, df_cells_female_5dose$log2FoldChange.y)

# PDF
pdf(file = here("output", "placenta", "cells_female_dose_correlation.pdf"))
smoothScatter(df_cells_female_5dose$log2FoldChange.y ~ df_cells_female_5dose$log2FoldChange.x,
              colramp = colorRampPalette(c("white", "violet")),
              cex = 3, xlim = c(-5, 5), ylim = c(-5, 5),
              main = "5uM PFBS: HTR-8/SVneo vs Female Term Tissue",
              xlab = "Log2Fold Change (Cells 5uM)",
              ylab = "Log2Fold Change (Female Term Tissue 5uM)")
abline(v = 0, col = "gray40")
abline(h = 0, col = "gray40")
abline(lm(df_cells_female_5dose$log2FoldChange.y ~ df_cells_female_5dose$log2FoldChange.x), col = "red", lwd = 2)
dev.off()

# TIFF
tiff(filename = here("output", "placenta", "cells_female_dose_correlation.tiff"),
     width = 6, height = 6, units = "in", res = 300, compression = "lzw")
smoothScatter(df_cells_female_5dose$log2FoldChange.y ~ df_cells_female_5dose$log2FoldChange.x,
              colramp = colorRampPalette(c("white", "violet")),
              cex = 3, xlim = c(-5, 5), ylim = c(-5, 5),
              main = "5uM PFBS: HTR-8/SVneo vs Female Term Tissue",
              xlab = "Log2Fold Change (Cells 5uM)",
              ylab = "Log2Fold Change (Female Term Tissue 5uM)")
abline(v = 0, col = "gray40")
abline(h = 0, col = "gray40")
abline(lm(df_cells_female_5dose$log2FoldChange.y ~ df_cells_female_5dose$log2FoldChange.x), col = "red", lwd = 2)
dev.off()
```




