---
title: "Third_Trimester_20"
output:
  pdf_document: default
  html_document: default
date: "2025-06-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install DESeq2
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")

#install tidyverse (ggplot2 is in tidyverse)
#install.packages("tidyverse")

```
# Package Setup
```{r}
library( "DESeq2" )
library(tidyverse)
library(ggplot2)
```

# Load Feature Count Data 

```{r}
# Define the file path and set working directory 
setwd("C:/Users/Maggie Drelichman/Downloads/RNA seq PFBS/FeatureCounts PFBS")
```

```{r}
# third_1_20um 
third_1_20um_count_data <- read.csv('1_20um_counts.tabular', header = TRUE, sep = "\t")
colnames(third_1_20um_count_data)[2] = "third_1_20um"

# third_1_Control 

third_1_Control_count_data <- read.csv('1_Control_counts.tabular', header = TRUE, sep = "\t")
colnames(third_1_Control_count_data)[2] = "third_1_Control"

# third_2_20um 

third_2_20um_count_data <- read.csv('2_20um_counts.tabular', header = TRUE, sep = "\t")
colnames(third_2_20um_count_data)[2] = "third_2_20um"

# third_2_Control 

third_2_Control_count_data <- read.csv('2_Control_counts.tabular', header = TRUE, sep = "\t")
colnames(third_2_Control_count_data)[2] = "third_2_Control"

# third_3_20um

third_3_20um_count_data <- read.csv('3_20um_counts.tabular', header = TRUE, sep = "\t")
colnames(third_3_20um_count_data)[2] = "third_3_20um"

# third 3_Control 

third_3_Control_count_data <- read.csv('3_Control_counts.tabular', header = TRUE, sep = "\t")
colnames(third_3_Control_count_data)[2] = "third_3_Control"
```

# Join all count data into one data frame by full_join & remove NA values with 0 
```{r}

#put all data frames into list 

df_list <- list(third_3_Control_count_data, third_3_20um_count_data, third_2_Control_count_data, third_2_20um_count_data, third_1_Control_count_data, third_1_20um_count_data)

library(purrr)

# merge all data frames into list 
count_data <- purrr::reduce(df_list, dplyr::full_join, by= 'Geneid')


# make the Geneid the row id 
count_data <- count_data %>% remove_rownames %>% column_to_rownames(var="Geneid")
```
# Not Actually Necessary? 
```{r}
#count_data[is.na(count_data)] <- 0
#head(count_data)
```
# Create Metadata
## Goal: Comparing "control" vs. "treated" across three donors (trials)
```{r}
SampleName <- c(colnames(count_data))
treatment <- c("control", "treated", "control", "treated", "control", "treated")
trial <- c("one","one","two", "two","three","three")
meta_data <- data.frame(SampleName, treatment, trial)
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="SampleName")
meta_data$treatment <- as.factor(meta_data$treatment)
meta_data$trial <- as.factor(meta_data$trial)

head(meta_data)

```
# Double check that the name in the Columns in the Counts Matrix = Nmaes in the Meta Data Frame 

```{r}

all(colnames(count_data) %in% rownames(meta_data))

all(colnames(count_data) == rownames(meta_data))
```

# Create DESeqDataSet object
# Model gene expression as a function treatment, while controllin for trial differences 
```{r}
design_formula <- ~ treatment + trial
dds <- DESeqDataSetFromMatrix(countData = count_data,
                              colData = meta_data,
                              design = design_formula)
#shows object elements
dim(dds)
```
# Pre-filtering: by removing rows in which there are very few reads
# filter any counts less than 200 
```{r}
keep <- rowSums(counts(dds)) >= 200
dds <- dds[keep,]
dds
head(dds)

resultsNames(dds)
```

# Set reference factor, set the reference to be the control
```{r}
dds$treatment <- relevel(dds$treatment, ref = 'control')
#By setting the reference level for trial "one", DESq2 now adjusts for baseline expression in "trial one" and compares "trial two" and "three" against it
#This adjustment removes confounding donor effects from the treatment comparison 
dds$trial <- relevel(dds$trial, ref = 'one')

resultsNames(dds)
```

# PCA plot
# Transform the count data using rlog
```{r}
rld.all <- rlog(dds)
```
# Specify the grouping variable for intgroup (replace 'group_variable' with your actual grouping variable)
#```{r}
#pdf(file = "C:/Users/Maggie Drelichman/Downloads/RNA seq PFBS/third_trimester_20.pdf")

#group_variables <- c("treatment","trial")
#DESeq2::plotPCA(rld.all, intgroup="treatment")
#DESeq2::plotPCA(rld.all, intgroup="trial")
#dev.off()

#```

```{r}
p1 <- plotPCA(rld.all, intgroup = "treatment") +
  ggtitle("PCA Plot: Treatment Groups (Control vs 20 µM PFBS)")
p2 <- plotPCA(rld.all, intgroup = "trial") +
  ggtitle("PCA Plot: Trial Groups (Donor Variation)")


ggsave("treatment_PCA.pdf", plot = p1)
ggsave("trial_PCA.pdf", plot = p2)
```
# Test 
```{r}
dds <- DESeq(dds) #testing

res.all <- results(dds, independentFiltering = FALSE)  # Extract results table

mcols(res.all)
summary(res.all,alpha=0.05) #viewing summary of DE genes

res.all$gene <- rownames(res.all) #adding actual columnn with gene names
res.all <- res.all [,c('gene','log2FoldChange','baseMean','lfcSE', 'stat','pvalue','padj')] #reordering columns
rownames(res.all)<-NULL #removing non-column row names (genes)
dim(res.all) #checking dataframe
head(res.all)#checking dataframe

log2foldchangeresults_file_path <- "C:/Users/Maggie Drelichman/Downloads/RNA seq PFBS/results_third_20.csv"
write.csv(as.data.frame(res.all), file = log2foldchangeresults_file_path, row.names = TRUE)

resultsNames(dds)
```
# Identifying Up and Down Regulated Genes 
```{r}
# add an additional column that identifies a gene as unregulated, downregulated, or unchanged
# note the choice of pvalue and log2FoldChange cutoff. 
res.all <- as.data.frame(res.all)
res.all <- res.all %>%
  mutate(
  Expression = case_when(log2FoldChange >= log(1) & padj <= 0.05 ~ "Up-regulated",
  log2FoldChange <= -log(1) & padj <= 0.05 ~ "Down-regulated",
  TRUE ~ "Unchanged")
  )
head(res.all)
```
# Which Genes are the Top 10 Most Variable?
```{r}
top <- 10
# we are getting the top 10 up and down regulated genes by filtering the column Up-regulated and Down-regulated and sorting by the adjusted p-value. 

top_genes <- bind_rows(
  res.all %>%
  filter(Expression == 'Up-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top),
  res.all %>%
  filter(Expression == 'Down-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top)
  )

# create a data frame just holding the top 10 genes

Top_Hits = head(arrange(res.all,pvalue),10)
Top_Hits
```

# Volcano Plots 
```{r}
library(RColorBrewer)
library(dplyr)

plot_volcano <- function(res, title = "Volcano Plot") {
  palette <- brewer.pal(4, "Reds")
  
  # Make sure rownames are stored in a column called 'gene'
  res <- as.data.frame(res)
  res$gene <- rownames(res)
  
  # Base volcano plot
  with(res, plot(log2FoldChange, -log10(pvalue), pch = 20, main = title, xlim = c(-4, 4),
                 ylim = c(0, 20), xlab = expression(Log[2]~Fold~Change), 
                 ylab = expression(-Log[10]~Pvalue), col = "grey"))
  
  with(subset(res, padj < 0.05), points(log2FoldChange, -log10(pvalue), pch = 20, col = palette[2]))
  with(subset(res, abs(log2FoldChange) > 1), points(log2FoldChange, -log10(pvalue), pch = 20, col = palette[3]))
  with(subset(res, padj < 0.05 & abs(log2FoldChange) > 1), points(log2FoldChange, -log10(pvalue), pch = 20, col = palette[4]))
  
  legend("topleft", legend = c("Not significant", "FDR < 0.05", "Abs(Log2FoldChange) > 1", 
                               "FDR < 0.05 & Abs(Log2FoldChange) > 1"), 
         col = c("grey", palette[2:4]), pch = 20)
  abline(h = -log10(0.05), col = "black", lty = 2)
  abline(v = c(-1, 1), col = "black", lty = 2)
  
  # Identify and label top 10 genes (by p-value)
  top_genes <- res %>%
    filter(padj < 0.05) %>%
    arrange(pvalue) %>%
    head(10)
  
  with(top_genes, text(log2FoldChange, -log10(pvalue), labels = gene, cex = 0.6, pos = 3))
}

plot_volcano(res.all, title = "Third Trimester Placenta: Control vs 20 µM PFBS")

pdf("C:/Users/Maggie Drelichman/Downloads/RNA seq PFBS/volcano_plot_20uM.pdf", width = 7, height = 6)
plot_volcano(res.all, title = "Third Trimester Placenta: Control vs 20 µM PFBS")
dev.off()
```
